### 2014-07-30 paper.tex
first commit

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}

\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

\lstdefinestyle{cxx}{ %
language=C++,
basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
% numbers=left,                   % where to put the line-numbers
% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
% numbersep=5pt,                  % how far the line-numbers are from the code
% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
commentstyle=\color{cyan},
stringstyle=\ttfamily\color{red},
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,                 % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
% morekeywords={char,double},
escapeinside={\%*}{*)},          % if you want to add a comment within your code
% morekeywords={*,...}          % if you want to add more keywords to the set
% morekeywords={char,double}    % if you want to add more keywords to the set
frameround=fttt
}

\lstdefinestyle{fortran}{ %
language=Fortran,
basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
% numbers=left,                   % where to put the line-numbers
% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
% numbersep=5pt,                  % how far the line-numbers are from the code
% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
commentstyle=\color{cyan},
stringstyle=\ttfamily\color{red},
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,                 % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
% morekeywords={char,double},
escapeinside={\%*}{*)},          % if you want to add a comment within your code
% morekeywords={*,...}          % if you want to add more keywords to the set
% morekeywords={char,double}    % if you want to add more keywords to the set
frameround=fttt
}

\begin{document}

%% \title{A Sample {\ttlit ACM} SIG Proceedings Paper in LaTeX
%% Format\titlenote{(Does NOT produce the permission block, copyright information nor page numbering). For use with ACM\_PROC\_ARTICLE-SP.CLS. Supported by ACM.}}
%% \subtitle{[Extended Abstract]
%% \titlenote{A full version of this paper is available as
%% \textit{Author's Guide to Preparing ACM SIG Proceedings Using
%% \LaTeX$2_\epsilon$\ and BibTeX} at
%% \texttt{www.acm.org/eaddress.htm}}}

%\conferenceinfo{EuroMPI}{2013 Madrid, Spain}

\title{
Efficient evaluation of threshold queries in a numerical simulation database
}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
\alignauthor Kalin Kanov \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{kalin@cs.jhu.edu}
\alignauthor Randal Burns \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University}\\
     \affaddr{Baltimore, MD 21218} \\
     \email{randal@cs.jhu.edu} 
\alignauthor Cristian C. Lalescu \\
     \affaddr{Department of Applied Mathematics and Statistics} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{clalesc1@jhu.edu}
}
%% \and % use '\and' if you need 'another row' of author names
%% % There's nothing stopping you putting the seventh, eighth, etc.
%% % author on the opening page (as the 'third row') but we ask,
%% % for aesthetic reasons that you place these 'additional authors'
%% % in the \additional authors block, viz.
%% \additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
%% email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
%% (The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
%% \date{30 July 1999}
%% % Just remember to make sure that the TOTAL number of authors
%% % is the number that will appear on the first page PLUS the
%% % number that will appear in the \additionalauthors section.

\maketitle

\begin{abstract}

\end{abstract}

%% % A category with the (minimum) three required fields
%\category{}{Extending Data Base Technology}
%% %A category including the fourth, optional field follows...
%% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

%\terms{Extending Database Technology}

\keywords{databases, data-intensive computing, threshold queries, turbulence} % NOT required for Proceedings

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at JHU, which have capacity of 1.1PB and xPB
respectively. These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. Subsequently, applications developed on top of these platforms have to make efficient use of all that they
have to offer. The tasks performed in a typical workflow are often data-intensive, but in a lot of case they can also be computationally intensive. 
This calls for the development of new methods for the efficient evaluation of the queries submitted to the system.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-k queries and similarity search in general
are important in many different disciplines \cite{TODO}. In this paper we introduce an efficient evaluation strategy for threshold
queries over time series datasets stored in a cluster of relational databases.

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries requires a kernel computation accessing several data points to be evaluated at each point on the grid.
This means that in order to evaluate such a query for a given time step all of the data for that time step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.

%Why is it interesting and important?
Thresholding allows scientists to obtain and examine the most intense regions and features in the dataset in the case of turbulence. 
These are often the locations that have the largest vorticity norms and have intense vortices or reconnection events. 
It is important that threshold queries are evaluated in an efficient manner, because often further subsequent examination
and analysis is required to understand the physics behind these intense events.

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures, clusters, etc. have only recently emerged and the systems that run on them often don't support
all of the functionality that scientists are interested in. In this paper we present a method for the efficient evaluation of 
threshold queries over numerical simulation data stored in a cluster of relational databases. We exploit the parallelism 
available in the cluster by means of data parallel execution of the computations and moreover introduce an application-aware
cache to store the query results.

%What are the key components of my approach and results? Also include any specific limitations.
- 3d and 4d clusters of highest vorticity locations.

- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.

- The introduction of an application-aware cache improves the running time of queries substantially if there is a cache hit and there is 
little downside to storing query results in the cache.

- The evaluation scales with the amount of parallelism available in the cluster.

-The evaluation is computationally limited and embarrassingly parallel, which makes it a good candidate for a GPU implementation.

%% * collection of statistics at runtime
%% * unable to traverse backwards in time
%% * stored in flat files
%% * difficulty reusing data or sharing
%% * example of JHU-TDC
%% * successful works of JHU-TDC

\section{Scientific Use Cases}
Max vorticity clustering.

\section{JHTDB}
The JHTDB store the entire space-time history of high-resolution fluid dynamics numerical simulations in
a cluster of relational databases.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls.
At present there are three datasets available publicly and one more available internally and to collaborators. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics, channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 150 TB.

The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
is running Windows Server 2008 and SQL Server 2008 R2. The data are partitioned spatially across 4 to 8
database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
curve to distribute the data across nodes and databases as well as to spatially subdivide each time step
into database atoms, which are of size $8^3$. Each such atom is indexed by the time step, which it belongs to and
by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
Queries to the data and derived fields, such as derivatives and filtered quantities are evaluated through
stored procedures or user-defined functions implemented in the Common Language Runtime (CLR) framework.

The Web-services are hosted on a front-end Web-server, which handles user requests and hosts the main Webpage portal.
The Web-server acts as a mediator sending the users' requests to the database nodes and initiating their distributed evaluation. 
It assembles the results and sends them back to the client

\section{Threshold Query Evaluation}
Application-aware cache that effectively caches not only data, but also computational effort. 
In the case of thresholding what is cached are the results of the computation. 
This means that the computation does not have to be repeated and the amount of data cached is several orders of magnitude smaller than the raw data.

First, check the cache. If data are available in the cache return them from there.
Make sure that the threshold stored in the cache is lower than the one specified.
If the threshold stored in the cache is higher or the data are not available perform the computation. 
After the results are sent back store the data in the cache.
If the cache does not have enough space to store the new data, delete the least recently used
cached entries to free up space.
This is all done within a transaction.

\section{Experimental Results}
-- measure the overhead of querying the cache first

-- measure time taken to perform the thresholding computation for an entire time step on a cold cache

-- measure time taken to evaluate the query after warming up the cache and polluting it afterward

-- compare with local evaluation

-- provide scaling measurements

\section{Related Work}
Asfalg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to
the time series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time series
stored in the database (usually one dimensional sequence of measurements) and the time series given in the query. In contrast, our approach focuses on
reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.

In a system called the tree cache Lopez et al. \cite{Lopez} make use of a small application-aware cache to reduce access time to large datasets stored on
disk. The tree cache stores individual octants of octree datasets and exploits application-specific information to determine which octants to cache and to 
perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache
we do not cache raw data objects, but rather query results. This substantially reduces the size of the cache and improves query performance.

\section{Conclusions and Future Work}

%% \section{The {\secit Body} of The Paper}
%% Typically, the body of a paper is organized
%% into a hierarchical structure, with numbered or unnumbered
%% headings for sections, subsections, sub-subsections, and even
%% smaller sections.  The command \texttt{{\char'134}section} that
%% precedes this paragraph is part of such a
%% hierarchy.\footnote{This is the second footnote.  It
%% starts a series of three footnotes that add nothing
%% informational, but just give an idea of how footnotes work
%% and look. It is a wordy one, just so you see
%% how a longish one plays out.} \LaTeX\ handles the numbering
%% and placement of these headings for you, when you use
%% the appropriate heading commands around the titles
%% of the headings.  If you want a sub-subsection or
%% smaller part to be unnumbered in your output, simply append an
%% asterisk to the command name.  Examples of both
%% numbered and unnumbered headings will appear throughout the
%% balance of this sample document.

%% Because the entire article is contained in
%% the \textbf{document} environment, you can indicate the
%% start of a new paragraph with a blank line in your
%% input file; that is why this sentence forms a separate paragraph.

%% \subsection{Type Changes and {\subsecit Special} Characters}
%% We have already seen several typeface changes in this sample.  You
%% can indicate italicized words or phrases in your text with
%% the command \texttt{{\char'134}textit}; emboldening with the
%% command \texttt{{\char'134}textbf}
%% and typewriter-style (for instance, for computer code) with
%% \texttt{{\char'134}texttt}.  But remember, you do not
%% have to indicate typestyle changes when such changes are
%% part of the \textit{structural} elements of your
%% article; for instance, the heading of this subsection will
%% be in a sans serif\footnote{A third footnote, here.
%% Let's make this a rather short one to
%% see how it looks.} typeface, but that is handled by the
%% document class file. Take care with the use
%% of\footnote{A fourth, and last, footnote.}
%% the curly braces in typeface changes; they mark
%% the beginning and end of
%% the text that is to be in the different typeface.

%% You can use whatever symbols, accented characters, or
%% non-English characters you need anywhere in your document;
%% you can find a complete list of what is
%% available in the \textit{\LaTeX\
%% User's Guide}\cite{Lamport:LaTeX}.

%% \subsection{Math Equations}
%% You may want to display math equations in three distinct styles:
%% inline, numbered or non-numbered display.  Each of
%% the three are discussed in the next sections.

%% \subsubsection{Inline (In-text) Equations}
%% A formula that appears in the running text is called an
%% inline or in-text formula.  It is produced by the
%% \textbf{math} environment, which can be
%% invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
%% construction or with the short form \texttt{\$. . .\$}. You
%% can use any of the symbols and structures,
%% from $\alpha$ to $\omega$, available in
%% \LaTeX\cite{Lamport:LaTeX}; this section will simply show a
%% few examples of in-text equations in context. Notice how
%% this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
%% set here in in-line math style, looks slightly different when
%% set in display style.  (See next section).

%% \subsubsection{Display Equations}
%% A numbered display equation -- one set off by vertical space
%% from the text and centered horizontally -- is produced
%% by the \textbf{equation} environment. An unnumbered display
%% equation is produced by the \textbf{displaymath} environment.

%% Again, in either environment, you can use any of the symbols
%% and structures available in \LaTeX; this section will just
%% give a couple of examples of display equations in context.
%% First, consider the equation, shown as an inline equation above:
%% \begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
%% Notice how it is formatted somewhat differently in
%% the \textbf{displaymath}
%% environment.  Now, we'll enter an unnumbered equation:
%% \begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
%% and follow it with another numbered equation:
%% \begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
%% just to demonstrate \LaTeX's able handling of numbering.

%% \subsection{Citations}
%% Citations to articles \cite{bowman:reasoning, clark:pct, braams:babel, herlihy:methodology},
%% conference
%% proceedings \cite{clark:pct} or books \cite{salas:calculus, Lamport:LaTeX} listed
%% in the Bibliography section of your
%% article will occur throughout the text of your article.
%% You should use BibTeX to automatically produce this bibliography;
%% you simply need to insert one of several citation commands with
%% a key of the item cited in the proper location in
%% the \texttt{.tex} file \cite{Lamport:LaTeX}.
%% The key is a short reference you invent to uniquely
%% identify each work; in this sample document, the key is
%% the first author's surname and a
%% word from the title.  This identifying key is included
%% with each item in the \texttt{.bib} file for your article.

%% The details of the construction of the \texttt{.bib} file
%% are beyond the scope of this sample document, but more
%% information can be found in the \textit{Author's Guide},
%% and exhaustive details in the \textit{\LaTeX\ User's
%% Guide}\cite{Lamport:LaTeX}.

%% This article shows only the plainest form
%% of the citation command, using \texttt{{\char'134}cite}.
%% This is what is stipulated in the SIGS style specifications.
%% No other citation format is endorsed.

%% \subsection{Tables}
%% Because tables cannot be split across pages, the best
%% placement for them is typically the top of the page
%% nearest their initial cite.  To
%% ensure this proper ``floating'' placement of tables, use the
%% environment \textbf{table} to enclose the table's contents and
%% the table caption.  The contents of the table itself must go
%% in the \textbf{tabular} environment, to
%% be aligned properly in rows and columns, with the desired
%% horizontal and vertical rules.  Again, detailed instructions
%% on \textbf{tabular} material
%% is found in the \textit{\LaTeX\ User's Guide}.

%% Immediately following this sentence is the point at which
%% Table 1 is included in the input file; compare the
%% placement of the table here with the table in the printed
%% dvi output of this document.

%% \begin{table}
%% \centering
%% \caption{Frequency of Special Characters}
%% \begin{tabular}{|c|c|l|} \hline
%% Non-English or Math&Frequency&Comments\\ \hline
%% \O & 1 in 1,000& For Swedish names\\ \hline
%% $\pi$ & 1 in 5& Common in math\\ \hline
%% \$ & 4 in 5 & Used in business\\ \hline
%% $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
%% \hline\end{tabular}
%% \end{table}

%% To set a wider table, which takes up the whole width of
%% the page's live area, use the environment
%% \textbf{table*} to enclose the table's contents and
%% the table caption.  As with a single-column table, this wide
%% table will ``float" to a location deemed more desirable.
%% Immediately following this sentence is the point at which
%% Table 2 is included in the input file; again, it is
%% instructive to compare the placement of the
%% table here with the table in the printed dvi
%% output of this document.

%% \begin{table*}
%% \centering
%% \caption{Some Typical Commands}
%% \begin{tabular}{|c|c|l|} \hline
%% Command&A Number&Comments\\ \hline
%% \texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
%% \texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
%% \texttt{{\char'134}table}& 300 & For tables\\ \hline
%% \texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
%% \end{table*}
%% % end the environment with {table*}, NOTE not {table}!

%% \subsection{Figures}
%% Like tables, figures cannot be split across pages; the
%% best placement for them
%% is typically the top or the bottom of the page nearest
%% their initial cite.  To ensure this proper ``floating'' placement
%% of figures, use the environment
%% \textbf{figure} to enclose the figure and its caption.

%% This sample document contains examples of \textbf{.eps}
%% and \textbf{.ps} files to be displayable with \LaTeX.  More
%% details on each of these is found in the \textit{Author's Guide}.

%% \begin{figure}
%% \centering
%% \epsfig{file=fly.eps}
%% \caption{A sample black and white graphic (.eps format).}
%% \end{figure}

%% \begin{figure}
%% \centering
%% \epsfig{file=fly.eps, height=1in, width=1in}
%% \caption{A sample black and white graphic (.eps format)
%% that has been resized with the \texttt{epsfig} command.}
%% \end{figure}

%% As was the case with tables, you may want a figure
%% that spans two columns.  To do this, and still to
%% ensure proper ``floating'' placement of tables, use the environment
%% \textbf{figure*} to enclose the figure and its caption.

%% Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
%% used; use
%% the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
%% commands as appropriate for the different file types.

%% \subsection{Theorem-like Constructs}
%% Other common constructs that may occur in your article are
%% the forms for logical constructs like theorems, axioms,
%% corollaries and proofs.  There are
%% two forms, one produced by the
%% command \texttt{{\char'134}newtheorem} and the
%% other by the command \texttt{{\char'134}newdef}; perhaps
%% the clearest and easiest way to distinguish them is
%% to compare the two in the output of this sample document:

%% This uses the \textbf{theorem} environment, created by
%% the\linebreak\texttt{{\char'134}newtheorem} command:
%% \newtheorem{theorem}{Theorem}
%% \begin{theorem}
%% Let $f$ be continuous on $[a,b]$.  If $G$ is
%% an antiderivative for $f$ on $[a,b]$, then
%% \begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
%% \end{theorem}

%% The other uses the \textbf{definition} environment, created
%% by the \texttt{{\char'134}newdef} command:
%% \newdef{definition}{Definition}
%% \begin{definition}
%% If $z$ is irrational, then by $e^z$ we mean the
%% unique number which has
%% logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
%% \end{definition}

%% \begin{figure}
%% \centering
%% \psfig{file=rosette.ps, height=1in, width=1in,}
%% \caption{A sample black and white graphic (.ps format) that has
%% been resized with the \texttt{psfig} command.}
%% \end{figure}

%% Two lists of constructs that use one of these
%% forms is given in the
%% \textit{Author's  Guidelines}.

%% \begin{figure*}
%% \centering
%% \epsfig{file=flies.eps}
%% \caption{A sample black and white graphic (.eps format)
%% that needs to span two columns of text.}
%% \end{figure*}
%% and don't forget to end the environment with
%% {figure*}, not {figure}!

%% There is one other similar construct environment, which is
%% already set up
%% for you; i.e. you must \textit{not} use
%% a \texttt{{\char'134}newdef} command to
%% create it: the \textbf{proof} environment.  Here
%% is a example of its use:
%% \begin{proof}
%% Suppose on the contrary there exists a real number $L$ such that
%% \begin{displaymath}
%% \lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
%% \end{displaymath}
%% Then
%% \begin{displaymath}
%% l=\lim_{x\rightarrow c} f(x)
%% = \lim_{x\rightarrow c}
%% \left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
%% = \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
%% \frac{f(x)}{g(x)} = 0\cdot L = 0,
%% \end{displaymath}
%% which contradicts our assumption that $l\neq 0$.
%% \end{proof}

%% Complete rules about using these environments and using the
%% two different creation commands are in the
%% \textit{Author's Guide}; please consult it for more
%% detailed instructions.  If you need to use another construct,
%% not listed therein, which you want to have the same
%% formatting as the Theorem
%% or the Definition\cite{salas:calculus} shown above,
%% use the \texttt{{\char'134}newtheorem} or the
%% \texttt{{\char'134}newdef} command,
%% respectively, to create it.

%% \subsection*{A {\secit Caveat} for the \TeX\ Expert}
%% Because you have just been given permission to
%% use the \texttt{{\char'134}newdef} command to create a
%% new form, you might think you can
%% use \TeX's \texttt{{\char'134}def} to create a
%% new command: \textit{Please refrain from doing this!}
%% Remember that your \LaTeX\ source code is primarily intended
%% to create camera-ready copy, but may be converted
%% to other forms -- e.g. HTML. If you inadvertently omit
%% some or all of the \texttt{{\char'134}def}s recompilation will
%% be, to say the least, problematic.

%% \section{Conclusions}
%% This paragraph will end the body of this sample document.
%% Remember that you might still have Acknowledgments or
%% Appendices; brief samples of these
%% follow.  There is still the Bibliography to deal with; and
%% we will make a disclaimer about that here: with the exception
%% of the reference to the \LaTeX\ book, the citations in
%% this paper are to articles which have nothing to
%% do with the present subject and are used as
%% examples only.
%% %\end{document}  % This is where a 'short' article might terminate

\section{Acknowledgment}

%% %ACKNOWLEDGMENTS are optional
%% \section{Acknowledgments}
%% This section is optional; it is a location for you
%% to acknowledge grants, funding, editing assistance and
%% what have you.  In the present case, for example, the
%% authors would like to thank Gerald Murray of ACM for
%% his help in codifying this \textit{Author's Guide}
%% and the \textbf{.cls} and \textbf{.tex} files that it describes.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{paper}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
%% \appendix
%% %Appendix A
%% \section{Headings in Appendices}
%% The rules about hierarchical headings discussed above for
%% the body of the article are different in the appendices.
%% In the \textbf{appendix} environment, the command
%% \textbf{section} is used to
%% indicate the start of each Appendix, with alphabetic order
%% designation (i.e. the first is A, the second B, etc.) and
%% a title (if you include one).  So, if you need
%% hierarchical structure
%% \textit{within} an Appendix, start with \textbf{subsection} as the
%% highest level. Here is an outline of the body of this
%% document in Appendix-appropriate form:
%% \subsection{Introduction}
%% \subsection{The Body of the Paper}
%% \subsubsection{Type Changes and  Special Characters}
%% \subsubsection{Math Equations}
%% \paragraph{Inline (In-text) Equations}
%% \paragraph{Display Equations}
%% \subsubsection{Citations}
%% \subsubsection{Tables}
%% \subsubsection{Figures}
%% \subsubsection{Theorem-like Constructs}
%% \subsubsection*{A Caveat for the \TeX\ Expert}
%% \subsection{Conclusions}
%% \subsection{Acknowledgments}
%% \subsection{Additional Authors}
%% This section is inserted by \LaTeX; you do not insert it.
%% You just add the names and information in the
%% \texttt{{\char'134}additionalauthors} command at the start
%% of the document.
%% \subsection{References}
%% Generated by bibtex from your ~.bib file.  Run latex,
%% then bibtex, then latex twice (to resolve references)
%% to create the ~.bbl file.  Insert that ~.bbl file into
%% the .tex source file and comment out
%% the command \texttt{{\char'134}thebibliography}.
%% % This next section command marks the start of
%% % Appendix B, and does not continue the present hierarchy
%% \section{More Help for the Hardy}
%% The acm\_proc\_article-sp document class file itself is chock-full of succinct
%% and helpful comments.  If you consider yourself a moderately
%% experienced to expert user of \LaTeX, you may find reading
%% it useful but please remember not to change it.
%% \balancecolumns
%% % That's all folks!
\end{document}

### 2014-08-12 paper.tex
Work on the intro. Added JHTDB and Thresholding sections.

@@
 \title{
-Efficient evaluation of threshold queries in a numerical simulation database
+Efficient evaluation of threshold queries of derived fields in a numerical simulation database
 }
@@
 to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
-fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at JHU, which have capacity of 1.1PB and xPB
+fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at Johns Hopkins, which have capacity 
+of 1.1PB and xPB
 respectively. These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
-capabilities with I/O and network bandwidth. Subsequently, applications developed on top of these platforms have to make efficient use of all that they
-have to offer. The tasks performed in a typical workflow are often data-intensive, but in a lot of case they can also be computationally intensive. 
-This calls for the development of new methods for the efficient evaluation of the queries submitted to the system.
+capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
+storage engines and usually have complex analysis routines built-in, which has largely been driven by the "move the computation to the data"
+paradigm. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the study of the
+particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation strategies 
+and methods for their execution. They have to operate on large datasets distributed across multiple nodes of a cluster of relational databases 
+and in order to reduce their running times they have to make efficient use of the cluster resources.
 
@@
 are important in many different disciplines \cite{TODO}. In this paper we introduce an efficient evaluation strategy for threshold
-queries over time series datasets stored in a cluster of relational databases.
+queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
+scalar field data stored in the database, but also performs thresholding of derived fields.
 
 %Why is it hard? (E.g., why do naive approaches fail?)
-The evaluation of threshold queries requires a kernel computation accessing several data points to be evaluated at each point on the grid.
-This means that in order to evaluate such a query for a given time step all of the data for that time step must be read and each data point 
+The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
+In order to evaluate such a query for a given time step all of the data for that time step must be read and each data point 
 will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
+It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, threshold 
+queries have to be evaluated on-demand from the stored data.
+Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
+shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
+computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.
 
@@
 It is important that threshold queries are evaluated in an efficient manner, because often further subsequent examination
-and analysis is required to understand the physics behind these intense events.
+and analysis is required to understand the physics that drive these intense events.
 
 %Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
-Data-intensive architectures, clusters, etc. have only recently emerged and the systems that run on them often don't support
+Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
+to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
+datasets. The virtual laboratories built on these systems often do not support
 all of the functionality that scientists are interested in. In this paper we present a method for the efficient evaluation of 
-threshold queries over numerical simulation data stored in a cluster of relational databases. We exploit the parallelism 
-available in the cluster by means of data parallel execution of the computations and moreover introduce an application-aware
-cache to store the query results.
+threshold queries over numerical simulation data stored in a cluster of relational databases.
 
 %What are the key components of my approach and results? Also include any specific limitations.
-- 3d and 4d clusters of highest vorticity locations.
-
-- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.
+We exploit the parallelism 
+available in the cluster by means of data-parallel execution of the computations and moreover introduce an application-aware
+cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data objects we cache
+the actual results of the computation, which leads to query performance improvement of up to X times.
 
@@
 
--The evaluation is computationally limited and embarrassingly parallel, which makes it a good candidate for a GPU implementation.
+- 3d and 4d clusters of highest vorticity locations.
 
-%% * collection of statistics at runtime
-%% * unable to traverse backwards in time
-%% * stored in flat files
-%% * difficulty reusing data or sharing
-%% * example of JHU-TDC
-%% * successful works of JHU-TDC
+- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.
+
+%-The evaluation is computationally limited and embarrassingly parallel, which makes it a good candidate for a GPU implementation.
 
@@
 \section{JHTDB}
-The JHTDB store the entire space-time history of high-resolution fluid dynamics numerical simulations in
-a cluster of relational databases.
+The Johns Hopkins Turbulence Databases (JHTDB), built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
+the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
+The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
 Access to the data is provided by means of Web-services and a variety of analysis functions have been
 implemented and can be executed through Web-service calls.
-At present there are three datasets available publicly and one more available internally and to collaborators. 
-The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics, channel 
+At present the service hosts four datasets, which are available publicly. 
+The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
 flow turbulence and homogenous buoyancy driven turbulence.
-The total amount of space occupied by the datasets is over 150 TB.
+The total amount of space occupied by the datasets is over 230 TB.
 
 The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
-is running Windows Server 2008 and SQL Server 2008 R2. The data are partitioned spatially across 4 to 8
+is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset are partitioned spatially across 4 to 8
 database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
@@
 
+The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
+differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
+have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
+top-k queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
+threshold and top-k queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-k queries is much
+smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse previously
+computed results makes the query results suitable to caching.
+
 \section{Threshold Query Evaluation}
-Application-aware cache that effectively caches not only data, but also computational effort. 
-In the case of thresholding what is cached are the results of the computation. 
-This means that the computation does not have to be repeated and the amount of data cached is several orders of magnitude smaller than the raw data.
 
-First, check the cache. If data are available in the cache return them from there.
-Make sure that the threshold stored in the cache is lower than the one specified.
-If the threshold stored in the cache is higher or the data are not available perform the computation. 
-After the results are sent back store the data in the cache.
-If the cache does not have enough space to store the new data, delete the least recently used
-cached entries to free up space.
-This is all done within a transaction.
+Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
+cached in an application-aware
+cache. The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.
+In most cases threshold queries operate over an entire time-step. Therefore, each such query is subdivided by the mediator into queries submitted to
+each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
+need to be requested from adjacent nodes. The data are read into memory and the particular field requested is computed at each of the locations on
+the grid. The norm or absolute value of the derived field at each location is compared against the specified threshold and if the value is higher it is 
+maintained along with
+the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low. The list of locations and values are then maintained in a 
+cache, which is local to the database node and indicates the function, time-step, spatial region and the threshold requested. We use a least recently
+used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
+avoid dirty-reads or inconstant view of the cache.
+
+The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
+for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
+answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
+preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
+queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
+a small set of data and do not need to perform any additional computation.
+
+%Application-aware cache that effectively caches not only data, but also computational effort. 
+%In the case of thresholding what is cached are the results of the computation. 
+%This means that the computation does not have to be repeated and the amount of data cached is several orders of magnitude smaller than the raw data.
+
+The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
+are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
+retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
+cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
+parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
+adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
+the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
+for the new records, space is freed up by removing the least recently used data. Again, reading from, updating or modifying the cache is done within a
+transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction, which
+provides for a higher degree of parallelism. Figure \ref{figCacheLookup} illustrates this process.
 
 \section{Experimental Results}
+We evaluate the developed method for the execution of threshold queries to large numerical simulation datasets with the goal of analyzing the 
+benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
+that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
+a threshold locally by the user.
+
+The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
+the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
+The database nodes are 2.33 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
+SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
+arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
+database file.
+
 -- measure the overhead of querying the cache first
@@
 \section{Related Work}
-Asfalg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
+A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
 queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to

### 2014-08-14 paper.tex
Work on experiments section.

@@
 
-\section{Scientific Use Cases}
+\section{Scientific Use Cases}\label{science_use_cases}
 Max vorticity clustering.
@@
 
--- measure the overhead of querying the cache first
+For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
+have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl. As described in section \ref{science_use_cases}
+thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
+the development of the most intense vortices that are observed. [TODO: Consider showing other fields as well; possibly thresholding the velocity by itself]
+
+-- measure the overhead of querying the cache first:
+Figure showing avg. execution times of queries for vorticity with a high threshold, medium threshold and low threshold. For each threshold value
+plot the time to simply do the I/O, the time the query takes without the presence of a cache (the difference is essentially the time taken to compute
+the vorticity field at each location on the grid) and the time the query takes with the inclusion of a cache. This shows that the overhead associated
+with querying and updating the cache is not significant. Consider including a figure for a representative time-step showing the PDF of the vorticity field
+to show how the vorticity values are distributed and to get an understanding for the different threshold levels.
 
@@
 
+-- provide scaling measurements
+
 -- compare with local evaluation
-
--- provide scaling measurements
 

### 2014-08-14 paper.tex
General updates and some related works.

@@
 
+\newcommand{\kk}[1]{{\color{blue}{\it KK: #1}}}
+
 \begin{document}
@@
 \begin{abstract}
+Large numerical simulations are frequently used to study many different physical phenomena. The datasets produced by these simulations are in the TB and
+even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
+The systems built on-top of these architectures, sometimes called virtual laboratories, do not always support the analysis functionality that scientists are 
+interested in. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical simulation datasets
+stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. Additionally, we introduce an application-aware 
+cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.
 
@@
 Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
-functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-k queries and similarity search in general
-are important in many different disciplines \cite{TODO}. In this paper we introduce an efficient evaluation strategy for threshold
+functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-$k$ queries and similarity search in general
+are important in many different disciplines. In this paper we introduce an efficient evaluation strategy for threshold
 queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
@@
 The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
-In order to evaluate such a query for a given time step all of the data for that time step must be read and each data point 
+The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
+In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
 will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
@@
 datasets. The virtual laboratories built on these systems often do not support
-all of the functionality that scientists are interested in. In this paper we present a method for the efficient evaluation of 
-threshold queries over numerical simulation data stored in a cluster of relational databases.
+all of the functionality that scientists are interested in. Database systems support rollup queries, including top-$k$ queries, but in most cases these
+queries apply only simple linear score functions on the attribute values of individual records. 
+We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
+stored in the database. Our method computes and operates on derived fields, which are derived by performing kernel computations over the array data 
+stored in the database.
+Our method makes effective use of the cluster resources and achieves high throughput and scalability.
 
@@
 We exploit the parallelism 
-available in the cluster by means of data-parallel execution of the computations and moreover introduce an application-aware
-cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data objects we cache
-the actual results of the computation, which leads to query performance improvement of up to X times.
+available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
+is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
+application-aware cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data 
+objects we cache the actual results of the computation, which leads to query performance improvement of up to X times. In that regard the cache
+that we use shares more similarities with the idea of semantic caching \cite{Ren}.
 
@@
 \section{JHTDB}
-The Johns Hopkins Turbulence Databases (JHTDB), built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
+The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
+Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
 the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
@@
 database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
-curve to distribute the data across nodes and databases as well as to spatially subdivide each time step
-into database atoms, which are of size $8^3$. Each such atom is indexed by the time step, which it belongs to and
+curve to distribute the data across nodes and databases as well as to spatially subdivide each time-step
+into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
 by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
@@
 
-The Web-services are hosted on a front-end Web-server, which handles user requests and hosts the main Webpage portal.
+The Web-services are hosted on a front-end Web-server, which handles user requests and hosts the main Web-page portal.
 The Web-server acts as a mediator sending the users' requests to the database nodes and initiating their distributed evaluation. 
@@
 have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
-top-k queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
-threshold and top-k queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-k queries is much
+top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
+threshold and top-$k$ queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-$k$ queries is much
 smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse previously
@@
 a small set of data and do not need to perform any additional computation.
-
-%Application-aware cache that effectively caches not only data, but also computational effort. 
-%In the case of thresholding what is cached are the results of the computation. 
-%This means that the computation does not have to be repeated and the amount of data cached is several orders of magnitude smaller than the raw data.
+\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
+and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
+computation accessing multiple data points in a kernel around the target location.}
 
@@
 for the new records, space is freed up by removing the least recently used data.  Again, reading from, updating or modifying the cache is done within a
-transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction, which
-provides for a higher degree of parallelism. Figure \ref{figCacheLookup} illustrates this process.
+transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
+provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{figCacheLookup} illustrates this process.
 
@@
 thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
-the development of the most intense vortices that are observed. [TODO: Consider showing other fields as well; possibly thresholding the velocity by itself]
+the development of the most intense vortices that are observed. \kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
 
@@
 
--- measure time taken to perform the thresholding computation for an entire time step on a cold cache
+-- measure time taken to perform the thresholding computation for an entire time-step on a cold cache
 
@@
 \section{Related Work}
+The processing of top-$k$ queries has been studied extensively in the context of distributed and relational database systems. A survey of different techniques
+in the case of centralized processing is given in \cite{Ilyas}. In the case of distributed processing different approaches focus on horizontally 
+\cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deals with array data stored in
+a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the atribute values of
+individual records. In addition to top-$k$ queries we are also interested in threshold queries, where the result set includes all locations and values that are
+above a user prescribed threshold.
+
+Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
+idea of semantic caching \cite{Ren} and make use of brach caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
+The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
+at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
+simple linear score functions.
+
 A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold

### 2014-08-26 paper.tex
Adding some figures. Work on the experiments section.

@@
 \begin{abstract}
-Large numerical simulations are frequently used to study many different physical phenomena. The datasets produced by these simulations are in the TB and
+Large numerical simulations are commonplace in the study of many different physical phenomena. 
+The datasets produced by these simulations are in the TB and
 even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
-The systems built on-top of these architectures, sometimes called virtual laboratories, do not always support the analysis functionality that scientists are 
-interested in. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical simulation datasets
+The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
+interested in out of the box. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
+simulation datasets
 stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. Additionally, we introduce an application-aware 
@@
 capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
-storage engines and usually have complex analysis routines built-in, which has largely been driven by the "move the computation to the data"
+storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
 paradigm. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the study of the
@@
 the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
-for the new records, space is freed up by removing the least recently used data.  Again, reading from, updating or modifying the cache is done within a
+for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
 transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
@@
 the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
-The database nodes are 2.33 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
+The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
 SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
@@
 For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
-have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl. As described in section \ref{science_use_cases}
+have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
+\begin{align}
+\vec{\omega} = \vec{\nabla} \times \vec{v}.
+\label{eq:curl}
+\end{align}
+We use finite differencing methods of
+different orders for the evaluation of the curl. As described in section \ref{science_use_cases}
 thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
@@
 
--- measure the overhead of querying the cache first:
-Figure showing avg. execution times of queries for vorticity with a high threshold, medium threshold and low threshold. For each threshold value
-plot the time to simply do the I/O, the time the query takes without the presence of a cache (the difference is essentially the time taken to compute
-the vorticity field at each location on the grid) and the time the query takes with the inclusion of a cache. This shows that the overhead associated
-with querying and updating the cache is not significant. Consider including a figure for a representative time-step showing the PDF of the vorticity field
-to show how the vorticity values are distributed and to get an understanding for the different threshold levels.
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{Figures/cache_overhead.pdf}
+\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
+\label{fig:cache_overhead}
+\end{figure}
+
+The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
+stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
+compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
+For these experiments we requested
+the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only ~4,300 points (or 0.0004\%
+of all points) were above the threshold. For the second set a medium threshold was chosen and ~87,000 points (or 0.0081\% of all points) were above the
+threshold. Finally, a low threshold was chosen for the last set with ~900,000 points (or 0.0847\% of all points) above the threshold. For each set a random
+time-step was chosen and the queries were run against that time-step. The execution times were averaged over 10 runs. We utilized 4 threads per database
+node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
+the margin of error.
+\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
+to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{Figures/io_comparison.pdf}
+\caption{Execution time for threshold queries evaluated utilizing different number of
+threads per server compared with the time taken to perform the I/O only.}
+\label{fig:io_comparison}
+\end{figure}
 
@@
 
--- provide scaling measurements
+The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we issued queries
+with the same threshold values as in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
+the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
+compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
+Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.
+
+We observer a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.3 times when going 
+to 4 threads and little speedup is 
+observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
+processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
+the large variance that is seen in the running time of the queries. Additionally, SQL
+Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation data atoms along the boundary of
+each region will be request by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to running a single
+thread.
+\kk{Should we have scale-out experiments as well?}
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{Figures/scaleup.pdf}
+\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
+threads per server showcasing the vertical scalability of the implementation.}
+\label{fig:scaleup}
+\end{figure}
 

### 2014-08-26 paper.tex
Adding a diagram showcasing how a cache lookup is performed.

@@
 simulation datasets
-stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. Additionally, we introduce an application-aware 
+stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
 cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
+\caption{Cache lookup.}
+\label{fig:cache_diagram}
+\end{figure*}
+
 \section{Threshold Query Evaluation}
@@
 transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
-provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{figCacheLookup} illustrates this process.
+provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{fig:cache_diagram} illustrates 
+this process.
 
@@
 For these experiments we requested
-the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only ~4,300 points (or 0.0004\%
-of all points) were above the threshold. For the second set a medium threshold was chosen and ~87,000 points (or 0.0081\% of all points) were above the
-threshold. Finally, a low threshold was chosen for the last set with ~900,000 points (or 0.0847\% of all points) above the threshold. For each set a random
-time-step was chosen and the queries were run against that time-step. The execution times were averaged over 10 runs. We utilized 4 threads per database
+the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
+(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
+(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
+(or 0.0847\% of all points) above the threshold. For each set a random
+time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
+The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
+We utilized 4 threads per database
 node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
-the margin of error.
+the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
 \kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
@@
 \centering
-\includegraphics[width=3.5in]{Figures/io_comparison.pdf}
-\caption{Execution time for threshold queries evaluated utilizing different number of
-threads per server compared with the time taken to perform the I/O only.}
-\label{fig:io_comparison}
+\includegraphics[width=3.5in]{Figures/scaleup.pdf}
+\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
+threads per server showcasing the vertical scalability of the implementation.}
+\label{fig:scaleup}
 \end{figure}
@@
 
-The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we issued queries
-with the same threshold values as in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
+The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
+queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
 the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
@@
 the large variance that is seen in the running time of the queries. Additionally, SQL
-Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation data atoms along the boundary of
-each region will be request by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to running a single
-thread.
+Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
+the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to
+running a single thread.
 \kk{Should we have scale-out experiments as well?}
+
+Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
+the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
+observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieve from the
+same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
+I/O only with a single thread.
 
@@
 \centering
-\includegraphics[width=3.5in]{Figures/scaleup.pdf}
-\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
-threads per server showcasing the vertical scalability of the implementation.}
-\label{fig:scaleup}
+\includegraphics[width=3.5in]{Figures/io_comparison.pdf}
+\caption{Execution time for threshold queries evaluated utilizing different number of
+threads per server compared with the time taken to perform the I/O only.}
+\label{fig:io_comparison}
 \end{figure}

### 2014-08-28 paper.tex
Some changes to the figures.

@@
 will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
-It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, threshold 
-queries have to be evaluated on-demand from the stored data.
+It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
+and threshold queries on them have to be evaluated on-demand from the stored data.
 Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
@@
 We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
-stored in the database. Our method computes and operates on derived fields, which are derived by performing kernel computations over the array data 
+stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the array data 
 stored in the database.
@@
 is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
-application-aware cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data 
-objects we cache the actual results of the computation, which leads to query performance improvement of up to X times. In that regard the cache
-that we use shares more similarities with the idea of semantic caching \cite{Ren}.
-
-- The introduction of an application-aware cache improves the running time of queries substantially if there is a cache hit and there is 
-little downside to storing query results in the cache.
-
-- The evaluation scales with the amount of parallelism available in the cluster.
+application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
+on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
+we cache the actual results of the computation, which leads to query performance improvement of up to X times.
 
 - 3d and 4d clusters of highest vorticity locations.
-
-- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.
 
@@
 a small set of data and do not need to perform any additional computation.
-\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
-and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
-computation accessing multiple data points in a kernel around the target location.}
+%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
+%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
+%computation accessing multiple data points in a kernel around the target location.}
 
@@
 We utilized 4 threads per database
-node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
+node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 3\% and within
 the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
@@
 
-We observer a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.3 times when going 
+We observer nearly a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.4 times when going 
 to 4 threads and little speedup is 
@@
 Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
-the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to
+the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.6 times faster when compared to
 running a single thread.
@@
 Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
-idea of semantic caching \cite{Ren} and make use of brach caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
+idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
 The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
@@
 \section{Conclusions and Future Work}
+
+- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.
 

### 2014-08-28 paper.tex
Some work on the experiments section. Included a table with running times for queries with a cache hit.

@@
 \usepackage[usenames,dvipsnames]{color}
+\usepackage{multirow}
+\usepackage{array}
 %\usepackage{natbib}% bold math
@@
 a small set of data and do not need to perform any additional computation.
+
 %\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
@@
 %computation accessing multiple data points in a kernel around the target location.}
+
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+The space required to cache the maximum number of points per time-step including the index space and database overhead is $\sim$40MB. 
+Therefore for a dataset containing 
+1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a derived field for the entire dataset will
+require $\sim$5TB.
 
@@
 a threshold locally by the user.
+
+\begin{table}
+\centering
+\begin{tabular}{|c|c|c|c|} \hline
+\multirow{2}{0.8in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{2-4}
+ & No cache & \parbox[t]{0.7in}{With cache (miss)} & \parbox[t]{0.7in}{With cache (hit)} \\ \hline
+4247 & 97.1 & 100.2 & 0.5\\ \hline
+86580 & 113.7 & 115.9 & 1.2\\ \hline
+909274 & 111.6 & 115.0 & 9.1\\ \hline
+\end{tabular}
+\caption{Effectiveness of caching.}
+\label{tab:cache_effectiveness}
+\end{table}
 
@@
 time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
-The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
+The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
+Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
+have to be evaluated from the raw data. 
+The execution times were averaged over 10 runs.
 We utilized 4 threads per database
-node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 3\% and within
-the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
+node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
+less than 3\% and within
+the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
+time that can be observed is due to the larger result set that has to be returned to the user.
 \kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
@@
 
--- measure time taken to perform the thresholding computation for an entire time-step on a cold cache
-
--- measure time taken to evaluate the query after warming up the cache and polluting it afterward
+Table \ref{tab:cache_effectiveness} shows the effectiveness of the cache in the case of a cache hit. It includes the data from Fig. \ref{fig:cache_overhead}
+for the cases of execution in the absence of a cache and for cache misses. In the last column we show the running time of queries for which there is a 
+cache hit. To simulate this scenario we first warm up the cache by submitting the same set of threshold queries of the vorticity field as before. We then submit
+several more unrelated queries with different time-steps and threshold values in order to pollute the cache. Finally, we issue the original set of queries and 
+measure their running times. As we can see from these results the introduction of an application-aware cache for the results of threshold queries yields over
+an order of magnitude improvement for queries that hit the cache even for queries that have a low threshold and large result sets. Let us focus on the query
+with low threshold, which returns $\sim$900,000 points. Given that the number of points that can be returned from a threshold query is limited to 1,000,000
+it is likely that all subsequent queries to this time-step will result in a cache hit as their threshold is likely to be equal or higher than the cached one.
 
 The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
-queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
+queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
+time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
+The computations for all of 
 the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
@@
 observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
-processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
-the large variance that is seen in the running time of the queries. Additionally, SQL
+processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. Additionally, SQL
 Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
@@
 the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
-observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieve from the
+observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieved from the
 same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
@@
 
-- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.
+- Lays the ground work for the creation of a landmark database, which could store the locations of the highest vorticity regions in the dataset.
 

### 2014-09-02 paper.tex
Reorganized "Threshold Query Evaluation" section

@@
 Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
-cached in an application-aware
-cache. The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.
-In most cases threshold queries operate over an entire time-step. Therefore, each such query is subdivided by the mediator into queries submitted to
+cached in an application-aware cache. The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the 
+data across the nodes of the cluster.
+
+{\bf Derived fields computation}. The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
+queries that are of biggest interest to science users aim to obtain all locations where the values of a derived field are above a given threshold. Thus, the
+derived field in question has to be computed from the raw data first. For most derived fields of interest this computation has local support. It has an 
+associated localized kernel of computation around each grid node. Therefore, the value of the derived field at each grid node depends on the value of the
+stored field at the grid locations, which are part of the kernel of computation.
+
+{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Therefore, each such query is subdivided by 
+the mediator into queries submitted to
 each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
 need to be requested from adjacent nodes. The data are read into memory and the particular field requested is computed at each of the locations on
-the grid. The norm or absolute value of the derived field at each location is compared against the specified threshold and if the value is higher it is 
-maintained along with
+the grid. The same strategy applies when utilizing multiple threads per node. The norm or absolute value of the derived field at each location is compared 
+against the specified threshold and if the value is higher it is maintained along with
 the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
-the entire time-step or a significant fraction of it for queries with thresholds that are set too low. The list of locations and values are then maintained in a 
-cache, which is local to the database node and indicates the function, time-step, spatial region and the threshold requested. We use a least recently
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low.
+
+{\bf Application-aware cache for query results}. The central part of the evaluation strategy for threshold queries that we have developed is the 
+introduction of an application-aware cache
+for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
+answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
+cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
 used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
-avoid dirty-reads or inconstant view of the cache.
-
-The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
-for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
-answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
+avoid dirty-reads or inconstant view of the cache. Caching the query results
 preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
@@
 
-The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
+{\bf Overall execution of threshold queries}. The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
 asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
@@
 cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
 adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against

### 2014-09-03 paper.tex
Added PDF of vorticity. Some reorganization.

@@
 The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
-interested in out of the box. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
+interested in out of the box. Moreover, data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
+locally and can take days or months to iterate over the entire dataset. 
+In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
 simulation datasets
 stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
-cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.
+cache for query results, which has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
 
@@
 of 1.1PB and xPB
-respectively. These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
+respectively. They have been developed with the goal of providing persistent storage and public access to world-class numerical simulation data. 
+These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
 capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
@@
 will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
-It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
-and threshold queries on them have to be evaluated on-demand from the stored data.
+Obtaining the derived field and thresholding locally requires not only the computation of the derived field over an entire time-step but also the transfer of
+a large amount of data most of which are subsequently discarded. 
+It is also impractical to materialize all possible derived fields with the raw data
+due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
+have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
+
 Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
@@
 on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
-we cache the actual results of the computation, which leads to query performance improvement of up to X times.
+we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
 
@@
 \section{Scientific Use Cases}\label{science_use_cases}
-Max vorticity clustering.
+%Max vorticity clustering.
+One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
+The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
+reconnection events associated with them. 
+Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
+and make subsequent analysis queries as needed in order to study these events.
+
+Similar to the vorticity there are other fields of interest, which can be thresholded. These include ...
 
@@
 
-{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Therefore, each such query is subdivided by 
+{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
 the mediator into queries submitted to
@@
 used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
-avoid dirty-reads or inconstant view of the cache. Caching the query results
+avoid dirty-reads or inconsistent view of the cache. Caching the query results
 preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
@@
 
-\begin{table}
+\begin{figure}
 \centering
-\begin{tabular}{|c|c|c|c|} \hline
-\multirow{2}{0.8in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{2-4}
- & No cache & \parbox[t]{0.7in}{With cache (miss)} & \parbox[t]{0.7in}{With cache (hit)} \\ \hline
-4247 & 97.1 & 100.2 & 0.5\\ \hline
-86580 & 113.7 & 115.9 & 1.2\\ \hline
-909274 & 111.6 & 115.0 & 9.1\\ \hline
-\end{tabular}
-\caption{Effectiveness of caching.}
-\label{tab:cache_effectiveness}
-\end{table}
+\includegraphics[width=3.35in]{Figures/PDF.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
 
+\subsection{Experimental Setup}
 The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
@@
 have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
+%\begin{align}
+%&\sum_{i_0=0}^{1} \sum_{i_1=0}^{1} ... \sum_{i_{d-1}=0}^{1} {(-1)}^{i_0 + i_1 + ...+ i_{d-1}}\cdot  \nonumber \\
+%&\qquad s(x_0(i_0), x_1(i_1), ..., x_{d-1}(i_{d-1}))
+%\label{eq:summed_volume}
+%\end{align}
 \begin{align}
-\vec{\omega} = \vec{\nabla} \times \vec{v}.
+\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
+\times \nonumber \left( v_x, v_y, v_z \right)\\
+&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
+\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
 \label{eq:curl}
@@
 We use finite differencing methods of
-different orders for the evaluation of the curl. As described in section \ref{science_use_cases}
+different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
+adjacent grid node values as follows:  
+%\begin{align}
+%\frac{df}{dx}
+%\label{eq:curl}
+%\end{align}
+\begin{align} 
+\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
+&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
+\end{align}
+where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction.
+To illustrate the distribution of the values for the norm of the vorticity field in the MHD dataset and to get an appreciation for the different threshold values
+used in the experiments we show a coarse view of the probability density function for the vorticity in Fig. \ref{fig:pdf}.  
+As described in section \ref{science_use_cases}
 thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
-the development of the most intense vortices that are observed. \kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
+the development of the most intense vortices that are observed. 
+%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
 
@@
 \centering
-\includegraphics[width=3.5in]{Figures/cache_overhead.pdf}
+\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
 \caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
@@
 
+\subsection{Evaluation of cache effectiveness}
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|c|c|} \hline
+\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
+& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
+80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
+60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
+44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
+\end{tabular}
+\caption{Effectiveness of caching.}
+\label{tab:cache_effectiveness}
+\end{table}
 The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
@@
 For these experiments we requested
-the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
-(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
-(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
+the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
+(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
+(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
 (or 0.0847\% of all points) above the threshold. For each set a random
@@
 time that can be observed is due to the larger result set that has to be returned to the user.
-\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
-to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}
 
@@
 \centering
-\includegraphics[width=3.5in]{Figures/scaleup.pdf}
+\includegraphics[width=3.35in]{Figures/scaleup.pdf}
 \caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
@@
 
+\subsection{Scaling and Distributed Evaluation}
 The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
@@
 \centering
-\includegraphics[width=3.5in]{Figures/io_comparison.pdf}
+\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
 \caption{Execution time for threshold queries evaluated utilizing different number of

### 2014-09-03 paper.tex
Work on Conclusions section.

@@
 
-- 3d and 4d clusters of highest vorticity locations.
-
 %-The evaluation is computationally limited and embarrassingly parallel, which makes it a good candidate for a GPU implementation.
@@
 \section{Scientific Use Cases}\label{science_use_cases}
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/PDF.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
+
 %Max vorticity clustering.
@@
 
+The vorticity is computed from the velocity field by taking its curl:
+\begin{align}
+\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
+\times \nonumber \left( v_x, v_y, v_z \right)\\
+&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
+\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
+\label{eq:curl}
+\end{align}
+We use finite differencing methods of
+different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
+adjacent grid node values as follows:  
+\begin{align} 
+\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
+&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
+\end{align}
+where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction. The partial derivatives
+along $y$ and $z$ are computed in the same fashion. 
+Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
+of how the values are distributed in the dataset as a whole.
+This coarse view of the data can be used by scientists to guide the selection of threshold values.
+
+- 3d and 4d clusters of highest vorticity locations.
+
 Similar to the vorticity there are other fields of interest, which can be thresholded. These include ...
@@
 \section{JHTDB}
+
+\begin{figure*}
+\centering
+\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
+\caption{Cache lookup.}
+\label{fig:cache_diagram}
+\end{figure*}
+
 The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
@@
 computed results makes the query results suitable to caching.
-
-\begin{figure*}
-\centering
-\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
-\caption{Cache lookup.}
-\label{fig:cache_diagram}
-\end{figure*}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
+\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
+\label{fig:cache_overhead}
+\end{figure}
+
 {\bf Overall execution of threshold queries}. The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
@@
 \centering
-\includegraphics[width=3.35in]{Figures/PDF.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
+\includegraphics[width=3.35in]{Figures/scaleup.pdf}
+\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
+threads per server showcasing the vertical scalability of the implementation.}
+\label{fig:scaleup}
 \end{figure}
@@
 For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
-have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
-%\begin{align}
-%&\sum_{i_0=0}^{1} \sum_{i_1=0}^{1} ... \sum_{i_{d-1}=0}^{1} {(-1)}^{i_0 + i_1 + ...+ i_{d-1}}\cdot  \nonumber \\
-%&\qquad s(x_0(i_0), x_1(i_1), ..., x_{d-1}(i_{d-1}))
-%\label{eq:summed_volume}
-%\end{align}
-\begin{align}
-\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
-\times \nonumber \left( v_x, v_y, v_z \right)\\
-&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
-\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
-\label{eq:curl}
-\end{align}
-We use finite differencing methods of
-different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
-adjacent grid node values as follows:  
-%\begin{align}
-%\frac{df}{dx}
-%\label{eq:curl}
-%\end{align}
-\begin{align} 
-\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
-&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
-\end{align}
-where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction.
-To illustrate the distribution of the values for the norm of the vorticity field in the MHD dataset and to get an appreciation for the different threshold values
-used in the experiments we show a coarse view of the probability density function for the vorticity in Fig. \ref{fig:pdf}.  
+have to be computed from the stored data. It defined as the curl of the velocity field.
 As described in section \ref{science_use_cases}
-thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
+thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
 the development of the most intense vortices that are observed. 
@@
 
+\subsection{Evaluation of cache effectiveness}
+
+The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
+stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
+compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
+For these experiments we requested
+the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Fig. \ref{fig:pdf}, which shows the distribution of values of
+the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
+For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
+(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
+(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
+(or 0.0847\% of all points) above the threshold. For each set a random
+time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.
+
+The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
+Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
+have to be evaluated from the raw data. 
+The execution times were averaged over 10 runs.
+We utilized 4 threads per database
+node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
+less than 3\% and within
+the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
+time that can be observed is due to the larger result set that has to be returned to the user.
+
 \begin{figure}
 \centering
-\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
-\label{fig:cache_overhead}
+\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
+\caption{Execution time for threshold queries evaluated utilizing different number of
+threads per server compared with the time taken to perform the I/O only.}
+\label{fig:io_comparison}
 \end{figure}
 
-\subsection{Evaluation of cache effectiveness}
 \begin{table}[h]
@@
 \end{table}
-The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
-stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
-compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
-For these experiments we requested
-the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
-(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
-(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
-(or 0.0847\% of all points) above the threshold. For each set a random
-time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
-The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
-Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
-have to be evaluated from the raw data. 
-The execution times were averaged over 10 runs.
-We utilized 4 threads per database
-node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
-less than 3\% and within
-the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
-time that can be observed is due to the larger result set that has to be returned to the user.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/scaleup.pdf}
-\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
-threads per server showcasing the vertical scalability of the implementation.}
-\label{fig:scaleup}
-\end{figure}
 
@@
 The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
-queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
+queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
+but with varying number of threads per node. Cache entries for the
 time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
@@
 I/O only with a single thread.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
-\caption{Execution time for threshold queries evaluated utilizing different number of
-threads per server compared with the time taken to perform the I/O only.}
-\label{fig:io_comparison}
-\end{figure}
 
@@
 
-- Lays the ground work for the creation of a landmark database, which could store the locations of the highest vorticity regions in the dataset.
+We have presented an efficient strategy for the evaluation of threshold queries of derived fields in large numerical simulation datasets. 
+The fields, which are thresholded are derived from the stored simulation data in a distributed data-parallel manner. The computations
+scale with the cluster resources and are performed on the database nodes, where the data are stored. This new capability allows researches to quickly
+obtain and focus on regions of special interest even if they lack the computing capabilities or data transfer rates necessary to examine entire time-steps
+or large parts of the entire dataset.
+
+We have introduced an application-aware cache for the query results. The cache adds minimal overhead during the evaluation of queries even if there is
+a cache miss and has modest storage requirements. Cache hits reduce query running times by over an order of magnitude.
+
+The introduction of an application-aware cache for query results lays the ground work for the creation of a landmark database. Such a database can store the 
+locations of the highest vorticity regions in the dataset or more broadly regions of interest and their associated statistics.
+
+The Web-services approach to archived numerical simulation datasets provides public access to high quality simulation data to anyone with an internet
+connection. The Web-services methods can be called from any modern programming language and we provide C, Fortran and Matlab client libraries for the
+JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by mens of a stored procedure
+or a user-defined function that has been implemented and deployed to handle these requests. This allows us to fine-tune the execution of these procedures
+and handle all requests transparently to the user. However, this approach also has drawbacks. Adding new functionality means adding to a long list of 
+Web-service calls and requires substantial implementation effort. In the case of threshold queries the stored procedure performing the evaluation must have
+an implementation for each derived field of interest even though the execution is handled by the same Web-service call.
+
+In the future we plan to develop
+declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
+implemented.
+Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}.
+In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which
+will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.
 
@@
 
+The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
+their insightful comments and suggestions. This work is supported in part by the National Science Foundation
+under Grants ....
+
 %% %ACKNOWLEDGMENTS are optional

### 2014-09-04 paper.tex
Moved JHTDB section. Work on Science use case section.

@@
 
-\section{Scientific Use Cases}\label{science_use_cases}
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/PDF.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
-\end{figure}
-
-%Max vorticity clustering.
-One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
-The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
-reconnection events associated with them. 
-Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
-and make subsequent analysis queries as needed in order to study these events.
-
-The vorticity is computed from the velocity field by taking its curl:
-\begin{align}
-\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
-\times \nonumber \left( v_x, v_y, v_z \right)\\
-&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
-\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
-\label{eq:curl}
-\end{align}
-We use finite differencing methods of
-different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
-adjacent grid node values as follows:  
-\begin{align} 
-\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
-&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
-\end{align}
-where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction. The partial derivatives
-along $y$ and $z$ are computed in the same fashion. 
-Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
-of how the values are distributed in the dataset as a whole.
-This coarse view of the data can be used by scientists to guide the selection of threshold values.
-
-- 3d and 4d clusters of highest vorticity locations.
-
-Similar to the vorticity there are other fields of interest, which can be thresholded. These include ...
-
 \section{JHTDB}
-
-\begin{figure*}
-\centering
-\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
-\caption{Cache lookup.}
-\label{fig:cache_diagram}
-\end{figure*}
 
@@
 
+\section{Scientific Use Cases}\label{science_use_cases}
+
+%Max vorticity clustering.
+One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
+The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
+reconnection events associated with them. 
+Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
+and make subsequent analysis queries as needed in order to study these events.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/PDF.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
+
+The vorticity is computed from the velocity field by taking its curl:
+\begin{align}
+\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
+\times \nonumber \left( v_x, v_y, v_z \right)\\
+&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
+\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
+\label{eq:curl}
+\end{align}
+We use finite differencing methods of
+different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
+adjacent grid node values as follows:  
+\begin{align} 
+\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
+&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
+\end{align}
+where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction. The partial derivatives
+along $y$ and $z$ are computed in the same fashion. 
+Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
+of how the values are distributed in the dataset as a whole.
+This coarse view of the data can be used by scientists to guide the selection of threshold values.
+
+- 3d and 4d clusters of highest vorticity locations.
+
+In addition to obtaining the regions of largest vorticity there is substantial interest in studying the regions with highest values for other fields, such as
+the second and third velocity gradient invariants ($Q$ and $R$). These invariants are scalar quantities whose values contain information about the 
+topology of the flow and the rates of vortex stretching and rotation. In MHD finding the locations with largest values for the electric current
+can lead to new insights into the development of the most intense reconnection events of magnetic field sheets in the simulated plasma. Similarly to
+the vorticity the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, which scientists would like to perform 
+threshold queries on certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.
+
 \section{Threshold Query Evaluation}
+
+\begin{figure*}
+\centering
+\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
+\caption{Cache lookup.}
+\label{fig:cache_diagram}
+\end{figure*}
 
@@
 
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|c|c|} \hline
+\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
+& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
+80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
+60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
+44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
+\end{tabular}
+\caption{Effectiveness of caching.}
+\label{tab:cache_effectiveness}
+\end{table}
+
 The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
@@
 \end{figure}
-
-\begin{table}[h]
-\centering
-\begin{tabular}{|c|c|c|c|c|} \hline
-\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
-& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
-80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
-60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
-44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
-\end{tabular}
-\caption{Effectiveness of caching.}
-\label{tab:cache_effectiveness}
-\end{table}
 

### 2014-09-06 paper.tex
Added scaleout experiments.

@@
 
+\usepackage{subfigure}
 \usepackage{listings}
@@
 of 1.1PB and xPB
-respectively. They have been developed with the goal of providing persistent storage and public access to world-class numerical simulation data. 
+respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
 These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
@@
 will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
-Obtaining the derived field and thresholding locally requires not only the computation of the derived field over an entire time-step but also the transfer of
-a large amount of data most of which are subsequently discarded. 
+Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
+but also the transfer of a large amount of data over the network most of which are subsequently discarded. One such local evaluation of a threshold
+query over an entire time-step would take over 20 hours in practice. Even when executed in parallel it would take months to iterate over the entire dataset.
 It is also impractical to materialize all possible derived fields with the raw data
@@
 have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
+This integrated approach evaluates a threshold query over an entire time-step in a few minutes as we show later.
 
@@
 The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
-is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset are partitioned spatially across 4 to 8
+is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset reside on a regular three-dimensional spatial grid with the
+exception of the channel flow data, which has an irregular $y$ dimension. The data are partitioned spatially across 4 to 8
 database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
@@
 The Web-server acts as a mediator sending the users' requests to the database nodes and initiating their distributed evaluation. 
-It assembles the results and sends them back to the client
+Each request is broken down into multiple parts based on the spatial layout of the data. Each part is asynchronously submitted for evaluation 
+to the database which stores the data needed for the evaluation.
+The Web-server assembles the results from the distributed computation and sends them back to the client.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/pdf.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
+
+\begin{figure*}
+\centering
+\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
+\caption{Cache lookup.}
+\label{fig:cache_diagram}
+\end{figure*}
 
@@
 top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
-threshold and top-$k$ queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-$k$ queries is much
-smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse previously
-computed results makes the query results suitable to caching.
+threshold and top-$k$ queries usually have to examine the entire data volume of a time-step. Furthermore, the data product of threshold and top-$k$ 
+queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
+previously computed results makes the query results suitable to caching.
 
@@
 and make subsequent analysis queries as needed in order to study these events.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/PDF.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
-\end{figure}
 
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
-\caption{Cache lookup.}
-\label{fig:cache_diagram}
-\end{figure*}
-
 Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
@@
 {\bf Derived fields computation}. The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
-queries that are of biggest interest to science users aim to obtain all locations where the values of a derived field are above a given threshold. Thus, the
+queries that are of biggest interest to science users produce all locations where the values of a derived field are above a given threshold. Thus, the
 derived field in question has to be computed from the raw data first. For most derived fields of interest this computation has local support. It has an 
 associated localized kernel of computation around each grid node. Therefore, the value of the derived field at each grid node depends on the value of the
-stored field at the grid locations, which are part of the kernel of computation.
+stored field at all of the grid locations, which are part of the kernel of computation.
 
@@
 each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
-need to be requested from adjacent nodes. The data are read into memory and the particular field requested is computed at each of the locations on
+need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. Such a band is
+needed on each of the sides of the box forming the domain of the computation.  
+The data are read into memory and the particular field requested is computed at each of the locations on
 the grid. The same strategy applies when utilizing multiple threads per node. The norm or absolute value of the derived field at each location is compared 
@@
 1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a derived field for the entire dataset will
-require $\sim$5TB.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
-\label{fig:cache_overhead}
-\end{figure}
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+require $\sim$5TB (15TB for vector fields).
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
+\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
+\label{fig:cache_overhead}
+\end{figure}
+
 \section{Experimental Results}
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=3.35in]{Figures/scaleup.pdf}
-\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
-threads per server showcasing the vertical scalability of the implementation.}
-\label{fig:scaleup}
-\end{figure}
+\subfigure[Vertical Scaling]{\includegraphics[width=3.35in]{Figures/scaleup.pdf}\label{fig:scaleup}}
+\subfigure[Horizontal Scaling]{\includegraphics[width=3.35in]{Figures/scaleout.pdf}\label{fig:scaleout}}
+\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The vertical scaling evaluation was performed 
+utilizing 1-8 threads per server on a 4-node cluster. The horizontal scaling evaluation was performed on 1 through 8 nodes.}
+\label{fig:scaling}
+\end{figure*}
 
@@
 For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
-have to be computed from the stored data. It defined as the curl of the velocity field.
+have to be computed from the stored data. It is defined as the curl of the velocity field.
 As described in section \ref{science_use_cases}
@@
 
-\begin{table}[h]
-\centering
-\begin{tabular}{|c|c|c|c|c|} \hline
-\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
-& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
-80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
-60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
-44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
-\end{tabular}
-\caption{Effectiveness of caching.}
-\label{tab:cache_effectiveness}
-\end{table}
-
 The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
@@
 
-\begin{figure}
+\begin{table}[h]
 \centering
-\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
-\caption{Execution time for threshold queries evaluated utilizing different number of
-threads per server compared with the time taken to perform the I/O only.}
-\label{fig:io_comparison}
-\end{figure}
+\begin{tabular}{|c|c|c|c|c|} \hline
+\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Average Running time (s.)}\\ \cline{3-5}
+& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
+80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
+60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
+44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
+\end{tabular}
+\caption{Effectiveness of caching.}
+\label{tab:cache_effectiveness}
+\end{table}
 
@@
 \subsection{Scaling and Distributed Evaluation}
-The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
+
+The data-parallel evaluation of threshold queries exhibits good vertical and nearly ideal horizontal scaling as shown in Fig. \ref{fig:scaling}. 
+For the scale-up (Fig. \ref{fig:scaleup}) experiments we used the same
 queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
@@
 compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
-Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.
-
-We observer nearly a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.4 times when going 
-to 4 threads and little speedup is 
+Each computation is independent and embarrassingly parallel. This allows us to make use of multiple threads per node and scale out to multiple database
+nodes.
+
+We observer nearly a two times speedup when going from a single thread per node to two threads per node (Fig. \ref{fig:scaleup}). 
+The speedup diminishes to 1.4 times when going to 4 threads and little speedup is 
 observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
@@
 running a single thread.
-\kk{Should we have scale-out experiments as well?}
+
+The scale-out experiments show a nearly perfect linear speedup as the evaluation is distributed to an increasing number of database nodes 
+(Fig. \ref{fig:scaleout}). For these experiments we issued queries with the same threshold levels as before to a cold cache. We utilized a single thread per
+database node to evaluate the horizontal scaling of the computation. The evaluation benefits not only from the additional computational resources with
+the addition of database nodes to the cluster but also from the increased memory size. The data needed for the computation of each derived field are
+read into memory and the larger memory size means that there is less contention with other system and application processes and it is less likely that
+virtual memory needs to be used. SQL Server also benefits from a larger buffer pool, which reduces the I/O time.
 
 Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
-the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
+the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear speedup is 
 observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieved from the
@@
 
--- compare with local evaluation
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
+\caption{Execution time for threshold queries evaluated utilizing different number of
+threads per server compared with the time taken to perform the I/O only.}
+\label{fig:io_comparison}
+\end{figure}
+
+So far we have presented the effectiveness of evaluating threshold queries of derived fields on the database cluster storing the raw simulation data. The 
+data-parallel computation of the derived fields allows us to evaluate a threshold query over
+an entire $1024^3$ time-step part of a 20TB dataset in less than two minutes. The introduction of an application-aware cache for the query results of these
+queries reduced this time to several seconds when there is a cache hit. In contrast, an evaluation of this functionality performed locally by the user takes
+over 20 hours. To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
+subregions of a time-step. This is necessary as requesting a derived field over an entire time-step will overload the network. Derived fields may have even
+more components than the scalar or vector field stored in the database. For example, the velocity gradient (needed for the computation of the vorticity) has 9
+components compared with the 3 components of the velocity. Given a single precision floating-point representation, this makes the velocity gradient of an
+entire time-step at least 36GB in size. A Web-service request will be much larger due to the overhead of wrapping the data in an xml format. After the
+field of interest is obtained locally the user has to threshold it to get the final result, which is reasonably fast, but discards most of the data that have been
+requested to yield a small in size result.
 
@@
 \cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deals with array data stored in
-a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the atribute values of
+a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the attribute values of
 individual records. In addition to top-$k$ queries we are also interested in threshold queries, where the result set includes all locations and values that are
-above a user prescribed threshold.
+above a user prescribed threshold. The threshold queries that we focus on operate over fields derived from the scalar or vector field simulation data through
+the evaluation of kernel computations at each data point.
 
@@
 perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache
-we do not cache raw data objects, but rather query results. This substantially reduces the size of the cache and improves query performance.
+we do not cache raw data objects, but rather query results. This substantially reduces the size of the cache and improves query performance. Additionally, 
+the cache that we introduce resides on disk rather than in memory, which greatly increases its potential size.
 
@@
 
-We have introduced an application-aware cache for the query results. The cache adds minimal overhead during the evaluation of queries even if there is
-a cache miss and has modest storage requirements. Cache hits reduce query running times by over an order of magnitude.
+We have introduced an application-aware cache for the query results of threshold queries. Cache hits reduce query running times by over an order of
+magnitude. The cache adds minimal overhead during the evaluation of queries even if there is
+a cache miss and has modest storage requirements. The cache is represented as a set of database tables and resides on disk rather than in memory. 
+Each database node has local cache tables, which allows for a bigger overall cache size.
 

### 2014-09-08 paper.tex
Discussion about most intense event

@@
 Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
-and make subsequent analysis queries as needed in order to study these events.
+and make subsequent analysis queries as needed in order to study these events. The relationship between different worms that connect and reconnect
+at those locations is of the most interest.
 
@@
 
-- 3d and 4d clusters of highest vorticity locations.
+Fig. \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
+in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
+from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. The timespan stored
+is equal to about one large eddy turnover time. It is not clear whether a different more intense event could emerge from an existing cluster if the timespan stored
+was larger. Fig. \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. There are several worms interacting in a complex
+way at the same time.
 

### 2014-09-08 paper.tex
Replaced cache_lookup figure with pseudocode.

--- 2014-09-06 paper.tex
@@
 \usepackage{subfigure}
+\usepackage{amsmath}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{varwidth}
 \usepackage{listings}
@@
 }
+
+\makeatletter
+\def\BState{\State\hskip-\ALG@thistlm}
+\makeatother
+
+\makeatletter
+\let\OldStatex\Statex
+\renewcommand{\Statex}[1][3]{%
+  \setlength\@tempdima{\algorithmicindent}%
+  \OldStatex\hskip\dimexpr#1\@tempdima\relax}
+\makeatother
 
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
-\caption{Cache lookup.}
-\label{fig:cache_diagram}
-\end{figure*}
+%\begin{figure*}
+%\centering
+%\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
+%\caption{Cache lookup.}
+%\label{fig:cache_diagram}
+%\end{figure*}
 
@@
 
-{\bf Overall execution of threshold queries}. The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
-asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
+\begin{algorithm}
+\caption{Get points above threshold using cache}\label{alg:cache_lookup}
+\begin{algorithmic}[1]
+\Require Dataset $d$, Field $f$, Timestep $t$, Threshold $k$, Query box $q = [x_l,y_l,z_l,x_u,y_u,z_u]$
+\Procedure{GetThreshold}{}
+\State $points \gets List()$
+\State $updateCache \gets false$
+\State $query \gets \text{SELECT * FROM cachedb..cacheInfo}$
+\Statex[4] $\text{WHERE dataset = {\it d} AND field = {\it f}}$
+\Statex[4] $\text{AND timestep = {\it t}}$
+\State $command \gets SqlCommand(query)$
+\State $reader \gets command.ExecuteReader()$
+\If {$reader.HasRows()$}
+\State $k_s \gets reader[``threshold"]$ \Comment{Stored threshold}
+\State $start \gets reader[``startIndex"]$
+\State $end \gets reader[``endIndex"]$
+\State $ordinal \gets reader[``ordinal"]$
+\If {$k \le k_s \And q \in \text{[start, end]}$}
+\State $\textit{query} \gets \text{SELECT * FROM cachedb..cacheData}$
+\Statex[4] $\text{WHERE cacheInfoOrdinal = {\it ordinal}}$
+\State $command \gets SqlCommand(query)$
+\State $reader \gets command.ExecuteReader()$
+\While{reader.Read()}
+\State $location \gets reader[``zindex"]$
+\State $norm \gets reader[``dataValue"]$
+\If {$norm \ge k \And location \in q$}
+\State $\textit{points.Add(new Point(location, norm))}$
+\EndIf
+\EndWhile
+\Else
+\State $updateCache \gets true$
+\EndIf
+\Else
+\State $updateCache \gets true$
+\EndIf
+\If {$updateCache$}
+\State $\text{Retrieve data covering {\it q} from DB.}$
+\ForAll{$p \in q$}
+\State{Compute $f$ at $p$.}
+\If{$\| f(p) \| \ge k$}
+\State $\textit{points.Add(new Point(p, $\| f(p) \|$))}$
+\EndIf
+\EndFor
+\State{Update cacheInfo and cacheData tables.}
+\EndIf
+\State{\Return $points$}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
+dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
+of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
+the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
+lookup.
+
+{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
+The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. First a cache lookup is performed.
+If the data for the requested field, time-step and spatial region
 are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
@@
 transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
-provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{fig:cache_diagram} illustrates 
-this process.
-
+provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
+
+\section{Experimental Results}
+We evaluate the developed method for the execution of threshold queries to large numerical simulation datasets with the goal of analyzing the 
+benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
+that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
+a threshold locally by the user.
 \begin{figure}
@@
 
-\section{Experimental Results}
-We evaluate the developed method for the execution of threshold queries to large numerical simulation datasets with the goal of analyzing the 
-benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
-that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
-a threshold locally by the user.
+\subsection{Experimental Setup}
+The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
+the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
+The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
+SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
+arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
+database file.
+
+For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
+have to be computed from the stored data. It is defined as the curl of the velocity field.
+As described in section \ref{science_use_cases}
+thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
+the development of the most intense vortices that are observed. 
+%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
+
+\subsection{Evaluation of cache effectiveness}
 
@@
 \end{figure*}
-
-\subsection{Experimental Setup}
-The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
-the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
-The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
-SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
-arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
-database file.
-
-For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
-have to be computed from the stored data. It is defined as the curl of the velocity field.
-As described in section \ref{science_use_cases}
-thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
-the development of the most intense vortices that are observed. 
-%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
-
-\subsection{Evaluation of cache effectiveness}
 

### 2014-09-09 paper.tex
Replaced cache_lookup figure with pseudocode.
add picture of most intense event
use NS vorticity histogram
also, I put in the axes for the most intense 3D cluster.

@@
 \centering
-\includegraphics[width=3.35in]{Figures/pdf.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\includegraphics[width=3.35in]{Figures/vorticity_histogram.pdf}
+\caption{Tail of the histogram of the norm $\omega$ of the vorticity field over the isotropic Navier-Stokes dataset. $\sqrt{\langle \omega ^2\rangle}$ is computed over the entire dataset.}
 \label{fig:pdf}
@@
 along $y$ and $z$ are computed in the same fashion. 
-Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
-of how the values are distributed in the dataset as a whole.
-This coarse view of the data can be used by scientists to guide the selection of threshold values.
+Fig. \ref{fig:pdf} shows the distribution of the large values of the norm of the vorticity for the isotropic Navier-Stokes dataset.
+This result can be used to estimate the number of points returned by a given threshold value (the integral of this curve must simply be divided by the total number of frames).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=3.35in]{Figures/most_intense_cluster.png}
+\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.
+Lines parallel to the axes are given for reference: red, green and blue correspond to $x$, $y$ and $z$, respectively;
+their lengths are the extents of the structure on each axis, which are approximately $0.55 \pi \times 0.40 \pi \times 0.17 \pi$.}
+\label{fig:max_vort_cluster}
+\end{figure}
 
@@
 in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
-from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. The timespan stored
-is equal to about one large eddy turnover time. It is not clear whether a different more intense event could emerge from an existing cluster if the timespan stored
-was larger. Fig. \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. There are several worms interacting in a complex
-way at the same time.
+from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. 
+The timespan stored is equal to about one large eddy turnover time. It is not clear whether a different more intense event could emerge from an existing
+cluster if the timespan stored was larger. Fig. \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
+There are several worms interacting in a complex way at the same time.
 
@@
 require $\sim$5TB (15TB for vector fields).
+
+The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
+dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
+of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
+the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
+lookup.
+
+{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
+The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
+First a cache lookup is performed.
+If the data for the requested field, time-step and spatial region
+are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
+retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
+cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
+adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
+the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
+for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
+transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
+provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
 
@@
 
-The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
-dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
-of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
-the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
-lookup.
-
-{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
-above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
-The mediator submits a query to each of the database nodes storing the raw data
-asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. First a cache lookup is performed.
-If the data for the requested field, time-step and spatial region
-are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
-retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
-cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
-adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
-the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
-for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
-transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
-provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
-
 \section{Experimental Results}
@@
 compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
+\textcolor{red}{Chichi: The following refers to the PDF made by Kalin; should we include that one as well or change this text?
+I just realized that the values are different since Kalin used the MHD dataset...
+Not only that, the statistics will be different as well, since the structures have different shapes.}
 For these experiments we requested

### 2014-09-09 paper.tex
SciDB, RasDaMan, SciQL references

@@
 an entire $1024^3$ time-step part of a 20TB dataset in less than two minutes. The introduction of an application-aware cache for the query results of these
-queries reduced this time to several seconds when there is a cache hit. In contrast, an evaluation of this functionality performed locally by the user takes
+queries reduces this time to several seconds when there is a cache hit. In contrast, an evaluation of this functionality performed locally by the user takes
 over 20 hours. To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
@@
 \section{Related Work}
+Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
+necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array simulation
+data in the JHTDB as a collection of binary large objects in a standard relational DBMS and perform the array manipulation tasks necessary at the application level.
+The systems that provide
+support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
+partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data is 
+handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB is 
+an array database system build from the ground up. It partitions arrays vertically into overlapping chunks. SciDB provides a declarative Array Query Language (AQL)
+and an Array Functional Language (AFL). Users can create arrays with named dimensions with AQL and make use of the functional operators defined in AFL, such
+as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
+within the MoneDB framework \cite{MonetDB}.
+
 The processing of top-$k$ queries has been studied extensively in the context of distributed and relational database systems. A survey of different techniques

### 2014-09-09 paper.tex
Switched back to MHD histogram.

@@
 \centering
-\includegraphics[width=3.35in]{Figures/vorticity_histogram.pdf}
-\caption{Tail of the histogram of the norm $\omega$ of the vorticity field over the isotropic Navier-Stokes dataset. $\sqrt{\langle \omega ^2\rangle}$ is computed over the entire dataset.}
+\includegraphics[width=3.35in]{Figures/pdf.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
 \label{fig:pdf}
@@
 along $y$ and $z$ are computed in the same fashion. 
-Fig. \ref{fig:pdf} shows the distribution of the large values of the norm of the vorticity for the isotropic Navier-Stokes dataset.
-This result can be used to estimate the number of points returned by a given threshold value (the integral of this curve must simply be divided by the total number of frames).
+Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
+of how the values are distributed in the dataset as a whole.
+This coarse view of the data can be used by scientists to guide the selection of threshold values.
 
@@
 lookup.
-
-{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
-above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
-The mediator submits a query to each of the database nodes storing the raw data
-asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
-First a cache lookup is performed.
-If the data for the requested field, time-step and spatial region
-are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
-retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
-cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
-adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
-the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
-for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
-transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
-provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
 
@@
 
+{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
+The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
+First a cache lookup is performed.
+If the data for the requested field, time-step and spatial region
+are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
+retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
+cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
+adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
+the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
+for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
+transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
+provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
+
 \section{Experimental Results}
@@
 compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
-\textcolor{red}{Chichi: The following refers to the PDF made by Kalin; should we include that one as well or change this text?
-I just realized that the values are different since Kalin used the MHD dataset...
-Not only that, the statistics will be different as well, since the structures have different shapes.}
 For these experiments we requested

### 2014-09-11 paper.tex
use alternate plot of most intense cluster

@@
 Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
-and make subsequent analysis queries as needed in order to study these events. The relationship between different worms that connect and reconnect
+and make subsequent analysis queries as needed in order to study these events. The relationship between different ``worms'' (see figure \ref{fig:max_vort_cluster}) that connect and reconnect
 at those locations is of the most interest.
@@
 
-\begin{figure}[h]
+\begin{figure*}[t]
 \centering
-\includegraphics[width=3.35in]{Figures/most_intense_cluster.png}
-\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.
-Lines parallel to the axes are given for reference: red, green and blue correspond to $x$, $y$ and $z$, respectively;
-their lengths are the extents of the structure on each axis, which are approximately $0.55 \pi \times 0.40 \pi \times 0.17 \pi$.}
+\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
+\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
 \label{fig:max_vort_cluster}
-\end{figure}
+\end{figure*}
 

### 2014-10-05 paper.tex
Added figures showing the breakdown of the exec. time. Some other changes throughout.

@@
 \begin{abstract}
-Large numerical simulations are commonplace in the study of many different physical phenomena. 
-The datasets produced by these simulations are in the TB and
-even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
-The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
-interested in out of the box. Moreover, data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
+%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
+%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
+%interested in out of the box. Moreover, d
+In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
+simulation datasets stored in a cluster of relational databases. 
+Such simulations are commonplace in the study of many different physical phenomena. 
+The datasets produced by these simulations are in the TB and even PB ranges. 
+Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
 locally and can take days or months to iterate over the entire dataset. 
-In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
-simulation datasets
-stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
-cache for query results, which has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
-
+The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution. 
+We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
+The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
 \end{abstract}
@@
 
-\keywords{databases, data-intensive computing, threshold queries, turbulence} % NOT required for Proceedings
+\keywords{scientific databases, data-intensive computing, threshold queries, turbulence} % NOT required for Proceedings
 
@@
 to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
-fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at Johns Hopkins, which have capacity 
-of 1.1PB and xPB
+fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
+of 1.1PB and 5PB
 respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
@@
 storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
-paradigm. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the study of the
-particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation strategies 
-and methods for their execution. They have to operate on large datasets distributed across multiple nodes of a cluster of relational databases 
-and in order to reduce their running times they have to make efficient use of the cluster resources.
+paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
+study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
+strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
+databases and in order to reduce their running times they have to make efficient use of the cluster resources.
 
@@
 Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
-but also the transfer of a large amount of data over the network most of which are subsequently discarded. One such local evaluation of a threshold
-query over an entire time-step would take over 20 hours in practice. Even when executed in parallel it would take months to iterate over the entire dataset.
-It is also impractical to materialize all possible derived fields with the raw data
+but also the transfer of a large amount of data over the network most of which are subsequently discarded. 
+One of our collaborators reported that such a local evaluation of a threshold
+query over an entire time-step would take over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
+would take months to iterate over the entire dataset. This highlighted the need for providing this important capability through an integrated approach,
+which performs the evaluation server-side.
+
+It is impractical to materialize all possible derived fields with the raw data
 due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
 have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
-This integrated approach evaluates a threshold query over an entire time-step in a few minutes as we show later.
+We will demonstrate that this integrated approach evaluates a threshold query over an entire time-step in a few minutes.
 
 Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
-shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
+shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. This is mainly because these application-independent 
+caches cannot exploit dataset-specific structure and application-level information. Moreover, even if the data are available in one or more of these
+application-independent caches the
 computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.
@@
 These are often the locations that have the largest vorticity norms and have intense vortices or reconnection events. 
+In magnetohydrodynamics, the locations of largest electric current are of great interest for similar reasons.
 It is important that threshold queries are evaluated in an efficient manner, because often further subsequent examination
@@
 to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
-datasets. The virtual laboratories built on these systems often do not support
-all of the functionality that scientists are interested in. Database systems support rollup queries, including top-$k$ queries, but in most cases these
-queries apply only simple linear score functions on the attribute values of individual records. 
+datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
+Relational database systems however often do not support
+all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
+capabilities have to be built-in through user-defined functions or stored procedures. 
+Database systems support rollup queries, including top-$k$ queries, but in most cases these
+queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
+Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning.
+This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
+Even approaches that aim to work with general score functions \cite{TODO} assume that the function operates on the attributes of a single record.
+In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
+are within the kernel of the computation.
+
+%What are the key components of my approach and results? Also include any specific limitations.
 We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
@@
 Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
-
-%What are the key components of my approach and results? Also include any specific limitations.
 We exploit the parallelism 
 available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
-is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
+is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
+system with an
 application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
-on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
+on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
 we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
+The cache that we describe is therefore a limited version of a semantic cache.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/pdf.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
 
@@
 
-\section{JHTDB}
+\section{JHTDB}\label{jhtdb}
 
 The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
-Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
+Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serve as a public virtual laboratory for
 the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
@@
 
-The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
+\begin{figure*}[t]
+\centering
+\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
+\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
+\label{fig:max_vort_cluster}
+\end{figure*}
+
+The database nodes are part of the GrayWulf \cite{Szalay} and Data-Scope \cite{DataScope} clusters. Each node
 is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset reside on a regular three-dimensional spatial grid with the
@@
 database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
-curve to distribute the data across nodes and databases as well as to spatially subdivide each time-step
+curve to distribute the data across nodes and databases. Each time-step is spatially subdivided
 into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
@@
 The Web-server assembles the results from the distributed computation and sends them back to the client.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/pdf.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
-\end{figure}
 
@@
 Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
-and make subsequent analysis queries as needed in order to study these events. The relationship between different ``worms'' (see figure \ref{fig:max_vort_cluster}) that connect and reconnect
+and make subsequent analysis queries as needed in order to study these events. The relationship between different ``worms'' (see Figure \ref{fig:max_vort_cluster}) that connect and reconnect
 at those locations is of the most interest.
@@
 along $y$ and $z$ are computed in the same fashion. 
-Fig. \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
+Figure \ref{fig:pdf} shows the distribution of the values of the norm of the vorticity field in the MHD dataset for a representative time-step. This is indicative 
 of how the values are distributed in the dataset as a whole.
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
-\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
-\label{fig:max_vort_cluster}
-\end{figure*}
-
-Fig. \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
+Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
 in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
 from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. 
-The timespan stored is equal to about one large eddy turnover time. It is not clear whether a different more intense event could emerge from an existing
-cluster if the timespan stored was larger. Fig. \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
+The timespan stored is equal to about one large eddy turnover time. 
+%It is not clear whether a different more intense event could emerge from an existing
+%cluster if the timespan stored was larger. 
+Figure \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
 There are several worms interacting in a complex way at the same time.
 
-In addition to obtaining the regions of largest vorticity there is substantial interest in studying the regions with highest values for other fields, such as
+In addition to obtaining the regions of largest vorticity, there is substantial interest in studying the regions with highest values for other fields, such as
 the second and third velocity gradient invariants ($Q$ and $R$). These invariants are scalar quantities whose values contain information about the 
-topology of the flow and the rates of vortex stretching and rotation. In MHD finding the locations with largest values for the electric current
+topology of the flow and the rates of vortex stretching and rotation. In MHD, finding the locations with largest values for the electric current
 can lead to new insights into the development of the most intense reconnection events of magnetic field sheets in the simulated plasma. Similarly to
-the vorticity the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, which scientists would like to perform 
+the vorticity, the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, which scientists would like to perform 
 threshold queries on certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.
@@
 Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
-cached in an application-aware cache. The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the 
-data across the nodes of the cluster.
-
-{\bf Derived fields computation}. The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
-queries that are of biggest interest to science users produce all locations where the values of a derived field are above a given threshold. Thus, the
-derived field in question has to be computed from the raw data first. For most derived fields of interest this computation has local support. It has an 
+cached in an application-aware cache, which is a limited version of a semantic cache. 
+The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.
+
+{\bf Derived fields computation:} The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
+queries of most interest to science users produce all locations where the values of a derived field are above a given threshold. Thus, the
+derived field in question has to be computed from the raw data first. For most derived fields of interest, this computation has local support. It has an 
 associated localized kernel of computation around each grid node. Therefore, the value of the derived field at each grid node depends on the value of the
@@
 
-{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
+{\bf Distributed data-parallel execution:} In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
 the mediator into queries submitted to
 each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
-need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. Such a band is
-needed on each of the sides of the box forming the domain of the computation.  
+need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. 
+Such a band is needed on each of the sides of the box forming the domain of the computation.  
 The data are read into memory and the particular field requested is computed at each of the locations on
-the grid. The same strategy applies when utilizing multiple threads per node. The norm or absolute value of the derived field at each location is compared 
+the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
 against the specified threshold and if the value is higher it is maintained along with
@@
 the entire time-step or a significant fraction of it for queries with thresholds that are set too low.
-
-{\bf Application-aware cache for query results}. The central part of the evaluation strategy for threshold queries that we have developed is the 
-introduction of an application-aware cache
-for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
-answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
-cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
-used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
-avoid dirty-reads or inconsistent view of the cache. Caching the query results
-preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
-queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
-a small set of data and do not need to perform any additional computation.
-
-%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
-%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
-%computation accessing multiple data points in a kernel around the target location.}
-
-The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
-The space required to cache the maximum number of points per time-step including the index space and database overhead is $\sim$40MB. 
-Therefore for a dataset containing 
-1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
-require $\sim$5TB (15TB for vector fields).
-
-The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
-dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
-of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
-the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
-lookup.
 
@@
 
-{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
-above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
+{\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 
+introduction of an application-aware cache
+for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
+answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
+cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
+used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
+avoid dirty-reads or an inconsistent view of the cache.
+
+Caching the query results
+preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
+and not the raw data fields.
+Thus, we do not have to derive the requested field from the raw data for 
+queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
+a small set of data and do not need to perform any additional computation.
+
+%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
+%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
+%computation accessing multiple data points in a kernel around the target location.}
+
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
+space and database overhead is $\sim$40MB. 
+Therefore for a dataset containing 
+1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+require $\sim$5TB (15TB for vector fields).
+
+The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
+dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
+of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
+the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
+lookup.
+
+{\bf Overall execution of threshold queries:} Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. 
 The mediator submits a query to each of the database nodes storing the raw data
@@
 cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
-adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
-the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
-for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
-transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
-provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested
+from adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is
+compared against the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not
+have enough space for the new records, space is freed up by removing the least recently used data across all quantities. Reading from, updating or
+modifying the cache is done within a transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the 
+cache for each transaction. This provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
 
@@
 benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
-that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
-a threshold locally by the user.
+that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than the
+user requesting the derived filed of interest from the database and evaluating the threshold locally.
+
+\subsection{Experimental Setup}
+The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
+the MHD dataset (Sec. \ref{jhtdb}) for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the
+Morton z-order. 
+The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
+SATA disks arranged as a set of four RAID-5 arrays. The database files are striped across the nodes and their associated disk
+arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
+database file.
+
 \begin{figure}
@@
 \includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
+\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
+absence of a cache.}
 \label{fig:cache_overhead}
 \end{figure}
-
-\subsection{Experimental Setup}
-The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
-the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
-The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
-SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
-arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
-database file.
 
@@
 \centering
-\subfigure[Vertical Scaling]{\includegraphics[width=3.35in]{Figures/scaleup.pdf}\label{fig:scaleup}}
-\subfigure[Horizontal Scaling]{\includegraphics[width=3.35in]{Figures/scaleout.pdf}\label{fig:scaleout}}
-\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The vertical scaling evaluation was performed 
-utilizing 1-8 threads per server on a 4-node cluster. The horizontal scaling evaluation was performed on 1 through 8 nodes.}
+\subfigure[Scale-up with multiple processes per node]{\includegraphics[width=3.35in]{Figures/scaleup.pdf}\label{fig:scaleup}}
+\subfigure[Scale-out to multiple nodes]{\includegraphics[width=3.35in]{Figures/scaleout.pdf}\label{fig:scaleout}}
+\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The scale-up evaluation was performed 
+utilizing 1-8 processes per server on a 4-node cluster. The scale-out evaluation was performed on 1 through 8 nodes.}
 \label{fig:scaling}
@@
 The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
-stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
-compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
-For these experiments we requested
-the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Fig. \ref{fig:pdf}, which shows the distribution of values of
+stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Figure
+\ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
+the cache first. For these experiments we requested
+the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
 the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
@@
 
+As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
+less than 3\% and within the margin of error.
 The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
-Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
-have to be evaluated from the raw data. 
+During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
+produce a cache miss and would have to be evaluated from the raw data. 
 The execution times were averaged over 10 runs.
-We utilized 4 threads per database
-node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
-less than 3\% and within
-the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
-time that can be observed is due to the larger result set that has to be returned to the user.
+We utilized 4 processes per database
+node for the evaluation of each query. Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
+misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
 
@@
 
-Table \ref{tab:cache_effectiveness} shows the effectiveness of the cache in the case of a cache hit. It includes the data from Fig. \ref{fig:cache_overhead}
-for the cases of execution in the absence of a cache and for cache misses. In the last column we show the running time of queries for which there is a 
-cache hit. To simulate this scenario we first warm up the cache by submitting the same set of threshold queries of the vorticity field as before. We then submit
+Cache hits reduce the running time of threshold queries by over an order of magnitude as shown in Figure \ref{fig:cache_overhead} and
+Table \ref{tab:cache_effectiveness}. This is because we do not have to compute the derived
+field requested from the raw data, which also eliminates the associated I/O. Only the cache entries need to be looked-up, which is substantially less data
+than the raw vector or scalar field data. For the queries with large result sets it is actually the network time taken to transfer the results to the user that 
+dominates the overall execution as opposed to the I/O or computation time as we show later.
+Cache hits are evaluated by first warming up the cache by submitting the same set of threshold queries of the vorticity field as before. We then submit
 several more unrelated queries with different time-steps and threshold values in order to pollute the cache. Finally, we issue the original set of queries and 
-measure their running times. As we can see from these results the introduction of an application-aware cache for the results of threshold queries yields over
-an order of magnitude improvement for queries that hit the cache even for queries that have a low threshold and large result sets. Let us focus on the query
+measure their running times. Let us focus on the query
 with low threshold, which returns $\sim$900,000 points. Given that the number of points that can be returned from a threshold query is limited to 1,000,000
@@
 \subsection{Scaling and Distributed Evaluation}
-
-The data-parallel evaluation of threshold queries exhibits good vertical and nearly ideal horizontal scaling as shown in Fig. \ref{fig:scaling}. 
+The evaluation of threshold queries of derived fields from the raw data is both I/O and computationally bound. These queries examine the entire data
+volume of a simulation time-step and are therefore good candidates for a data-parallel distributed evaluation. 
+Our data-parallel implementation exhibits good vertical and nearly ideal horizontal scaling as shown in Figure \ref{fig:scaling}. 
 For the scale-up (Fig. \ref{fig:scaleup}) experiments we used the same
-queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
-but with varying number of threads per node. Cache entries for the
+queries and threshold values as for the runs shown in Figure \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
+but with varying number of processes per node. Cache entries for the
 time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
@@
 compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
-Each computation is independent and embarrassingly parallel. This allows us to make use of multiple threads per node and scale out to multiple database
-nodes.
-
-We observer nearly a two times speedup when going from a single thread per node to two threads per node (Fig. \ref{fig:scaleup}). 
-The speedup diminishes to 1.4 times when going to 4 threads and little speedup is 
-observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
+Each computation is independent and embarrassingly parallel. This allows us to make use of multiple processes per node and scale out to
+multiple database nodes.
+
+We observer nearly a two times speedup when going from a single process per node to two processes per node (Fig. \ref{fig:scaleup}). 
+The speedup diminishes to 1.4 times when going to 4 processes and little speedup is 
+observed with 8 processes per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
 processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. Additionally, SQL
 Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
-the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.6 times faster when compared to
-running a single thread.
+the boundary of each region will be requested by multiple processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
+compared to running a single process.
 
 The scale-out experiments show a nearly perfect linear speedup as the evaluation is distributed to an increasing number of database nodes 
-(Fig. \ref{fig:scaleout}). For these experiments we issued queries with the same threshold levels as before to a cold cache. We utilized a single thread per
+(Fig. \ref{fig:scaleout}). For these experiments we issued queries with the same threshold levels as before to a cold cache. We utilized a single process per
 database node to evaluate the horizontal scaling of the computation. The evaluation benefits not only from the additional computational resources with
@@
 
-Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
-the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear speedup is 
-observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieved from the
-same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
-I/O only with a single thread.
+As expected we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
+Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
+node with
+the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. SQL Server already makes use of parallelism
+internally and the data have to be retrieved from the same set of disks. Nevertheless, the I/O time does decrease with additional processes, this is because
+the data reside in a partitioned table and the data in each partition are placed in a separate file on one of the disk arrays. Depending on how the data
+requests are scheduled in SQL Server this allows for the disks arrays to be driven in parallel. Additionally, with more processes per node the data can
+also be consumed faster. It is worth noting that the total running time for the queries evaluated with 4 or 8
+processes is about the same as the time it takes to perform the I/O only with a single process.
 
@@
 \caption{Execution time for threshold queries evaluated utilizing different number of
-threads per server compared with the time taken to perform the I/O only.}
+processes per server compared with the time taken to perform the I/O only.}
 \label{fig:io_comparison}
@@
 more components than the scalar or vector field stored in the database. For example, the velocity gradient (needed for the computation of the vorticity) has 9
-components compared with the 3 components of the velocity. Given a single precision floating-point representation, this makes the velocity gradient of an
+components compared with the 3 components of the velocity. Given a single-precision floating-point representation, this makes the velocity gradient of an
 entire time-step at least 36GB in size. A Web-service request will be much larger due to the overhead of wrapping the data in an xml format. After the
@@
 requested to yield a small in size result.
+
+\begin{figure*}
+\centering
+\subfigure[Vorticity]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown.pdf}\label{fig:vorticity_breakdown}}
+\subfigure[Q-criterion]{\includegraphics[width=2.2in]{Figures/q_breakdown.pdf}\label{fig:q_breakdown}}
+\subfigure[Magnetic field]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown.pdf}\label{fig:magnetic_breakdown}}
+\subfigure[Vorticity (cache hit)]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown_cachehit.pdf}\label{fig:vorticity_breakdown_cachehit}}
+\subfigure[Q-criterion (cache hit)]{\includegraphics[width=2.2in]{Figures/q_breakdown_cachehit.pdf}\label{fig:q_breakdown_cachehit}}
+\subfigure[Magnetic field (cache hit)]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown_cachehit.pdf}\label{fig:magnetic_breakdown_cachehit}}
+\caption{Breakdown of the execution time for threshold queries requesting different fields and at different threshold levels -- high, medium and low.}
+\label{fig:breakdown}
+\end{figure*}
+
+\subsection{Execution Time Breakdown}
+\kk{OR}
+\subsection{Evaluation of Additional Fields}
+The data-parallel evaluation of threshold queries shows stable execution time for different derived fields in addition to the different threshold levels and
+time-steps queried. The execution times depend on the complexity of the computation needed to evaluate the particular derived field requested. Figures 
+\ref{fig:vorticity_breakdown}, \ref{fig:q_breakdown}, \ref{fig:magnetic_breakdown} show a breakdown of the execution time of threshold queries of different
+derived fields, which are evaluated from the raw data and a cold cache using 4 processes per node on a 4 node cluster. Almost the entire time is spent
+performing the I/O and computation associated with the derived field requested.
+
+The vorticity field and the Q-criterion have similar I/O requirements as they
+have the same kernels of computation and both can be derived from the velocity gradient. The vorticity has 3 components and its computation only examines
+6 of the 9 components of the velocity gradient, which are also examined in pairs (see Eq. \ref{eq:curl}). On the other hand, even though the Q-criterion is a
+scalar value, it is computed through a non-linear combination of all 9 of the components of the velocity gradient. This means that the velocity gradient has
+to be computed at each grid location before the Q-criterion is evaluated, which is reflected in the increased computation time that we observer for the
+Q-criterion. The magnetic field is one of the raw fields of the magnetohydrodynamics dataset that are stored in the database. Therefore, there is no
+additional computation needed to derive it from the data, every data point has to be simply compared with the threshold level specified. This is why the 
+computation time is much smaller compared to the queries for the vorticity and the Q-criterion. The I/O time for the magnetic field is also smaller. This is
+because its kernel of computation is a single point and therefore there are no additional data along the boundary that have to be requested from adjacent 
+nodes. In that case all of the data needed are available locally.
+
+In all of these cases the time take to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
+time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase with the number of points in
+the result set.
+
+It is interesting to note that the time taken to perform a cache lookup is relatively small even in the case of a cache hit as can be seen in Figures 
+\ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit}, \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
+SSDs attached to each database node and retrieving the data is always done through a clustered index lookup. In the cases of a cache hit the majority of
+the time is spent simply transferring the results from the database nodes to the mediator and then back to the user. These times remain more or less
+constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
+Caching the results of threshold queries
+effectively preserves the I/O and computational effort spent during their initial evaluation and results in over an order of magnitude speedup for all the
+different fields requested as we can see in Figure \ref{fig:breakdown}.
 
@@
 handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB is 
-an array database system build from the ground up. It partitions arrays vertically into overlapping chunks. SciDB provides a declarative Array Query Language (AQL)
-and an Array Functional Language (AFL). Users can create arrays with named dimensions with AQL and make use of the functional operators defined in AFL, such
+an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping chunks.
+SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named dimensions with
+AQL and make use of the functional operators defined in AFL, such
 as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
-within the MoneDB framework \cite{MonetDB}.
+within the MonetDB framework \cite{MonetDB}.
 
@@
 a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the attribute values of
-individual records. In addition to top-$k$ queries we are also interested in threshold queries, where the result set includes all locations and values that are
-above a user prescribed threshold. The threshold queries that we focus on operate over fields derived from the scalar or vector field simulation data through
+individual records. In addition to top-$k$ queries, we are also interested in threshold queries, where the result set includes all locations and values that are
+above a user-prescribed threshold. The threshold queries that we focus on operate over fields derived from the scalar or vector field simulation data through
 the evaluation of kernel computations at each data point.
@@
 disk. The tree cache stores individual octants of octree datasets and exploits application-specific information to determine which octants to cache and to 
-perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache
+perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache,
 we do not cache raw data objects, but rather query results. This substantially reduces the size of the cache and improves query performance. Additionally, 
@@
 We have presented an efficient strategy for the evaluation of threshold queries of derived fields in large numerical simulation datasets. 
-The fields, which are thresholded are derived from the stored simulation data in a distributed data-parallel manner. The computations
+The thresholded fields are derived from the stored simulation data in a distributed data-parallel manner. The computations
 scale with the cluster resources and are performed on the database nodes, where the data are stored. This new capability allows researches to quickly
@@
 a cache miss and has modest storage requirements. The cache is represented as a set of database tables and resides on disk rather than in memory. 
-Each database node has local cache tables, which allows for a bigger overall cache size.
+Each database node has local cache tables, which allows the cache to scale-out as the cluster grows.
 

### 2014-10-06 paper.tex
Added JHTDB diagram.

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/jhtdb_diagram.pdf}
+\caption{Architectrue of the JHTDB.}
+\label{fig:jhtdb_diagram}
+\end{figure}
+
 %Why is it interesting and important?
@@
 This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
-Even approaches that aim to work with general score functions \cite{TODO} assume that the function operates on the attributes of a single record.
+Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
 In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
@@
 
-The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
-Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serve as a public virtual laboratory for
-the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
-The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
-Access to the data is provided by means of Web-services and a variety of analysis functions have been
-implemented and can be executed through Web-service calls.
-At present the service hosts four datasets, which are available publicly. 
-The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
-flow turbulence and homogenous buoyancy driven turbulence.
-The total amount of space occupied by the datasets is over 230 TB.
-
 \begin{figure*}[t]
@@
 \end{figure*}
+
+The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
+Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serve as a public
+virtual laboratory for
+the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
+The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
+Access to the data is provided by means of Web-services and a variety of analysis functions have been
+implemented and can be executed through Web-service calls (Fig. \ref{fig:jhtdb_diagram}).
+At present the service hosts four datasets, which are available publicly. 
+The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
+flow turbulence and homogenous buoyancy driven turbulence.
+The total amount of space occupied by the datasets is over 230 TB.
 
@@
 the entire time-step or a significant fraction of it for queries with thresholds that are set too low.
+
+{\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 
+introduction of an application-aware cache
+for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
+answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
+cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
+used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
+avoid dirty-reads or an inconsistent view of the cache.
+
+Caching the query results
+preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
+and not the raw data fields.
+Thus, we do not have to derive the requested field from the raw data for 
+queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
+a small set of data and do not need to perform any additional computation.
+
+%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
+%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
+%computation accessing multiple data points in a kernel around the target location.}
+
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
+space and database overhead is $\sim$40MB. 
+Therefore for a dataset containing 
+1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+require $\sim$5TB (15TB for vector fields).
 
@@
 
-{\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 
-introduction of an application-aware cache
-for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
-answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
-cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
-used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
-avoid dirty-reads or an inconsistent view of the cache.
-
-Caching the query results
-preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
-and not the raw data fields.
-Thus, we do not have to derive the requested field from the raw data for 
-queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
-a small set of data and do not need to perform any additional computation.
-
-%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
-%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
-%computation accessing multiple data points in a kernel around the target location.}
-
-The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
-Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
-space and database overhead is $\sim$40MB. 
-Therefore for a dataset containing 
-1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
-require $\sim$5TB (15TB for vector fields).
-
 The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
@@
 
+\begin{figure*}
+\centering
+\subfigure[Vorticity]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown.pdf}\label{fig:vorticity_breakdown}}
+\subfigure[Q-criterion]{\includegraphics[width=2.2in]{Figures/q_breakdown.pdf}\label{fig:q_breakdown}}
+\subfigure[Magnetic field]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown.pdf}\label{fig:magnetic_breakdown}}
+\subfigure[Vorticity (cache hit)]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown_cachehit.pdf}\label{fig:vorticity_breakdown_cachehit}}
+\subfigure[Q-criterion (cache hit)]{\includegraphics[width=2.2in]{Figures/q_breakdown_cachehit.pdf}\label{fig:q_breakdown_cachehit}}
+\subfigure[Magnetic field (cache hit)]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown_cachehit.pdf}\label{fig:magnetic_breakdown_cachehit}}
+\caption{Breakdown of the execution time for threshold queries requesting different fields and at different threshold levels -- high, medium and low.}
+\label{fig:breakdown}
+\end{figure*}
+
 So far we have presented the effectiveness of evaluating threshold queries of derived fields on the database cluster storing the raw simulation data. The 
@@
 requested to yield a small in size result.
-
-\begin{figure*}
-\centering
-\subfigure[Vorticity]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown.pdf}\label{fig:vorticity_breakdown}}
-\subfigure[Q-criterion]{\includegraphics[width=2.2in]{Figures/q_breakdown.pdf}\label{fig:q_breakdown}}
-\subfigure[Magnetic field]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown.pdf}\label{fig:magnetic_breakdown}}
-\subfigure[Vorticity (cache hit)]{\includegraphics[width=2.2in]{Figures/vorticity_breakdown_cachehit.pdf}\label{fig:vorticity_breakdown_cachehit}}
-\subfigure[Q-criterion (cache hit)]{\includegraphics[width=2.2in]{Figures/q_breakdown_cachehit.pdf}\label{fig:q_breakdown_cachehit}}
-\subfigure[Magnetic field (cache hit)]{\includegraphics[width=2.2in]{Figures/magnetic_breakdown_cachehit.pdf}\label{fig:magnetic_breakdown_cachehit}}
-\caption{Breakdown of the execution time for threshold queries requesting different fields and at different threshold levels -- high, medium and low.}
-\label{fig:breakdown}
-\end{figure*}
 

### 2014-10-07 paper.tex
Rearranged some figures. Cleaned up references.
Some minor fixes.

@@
 locally and can take days or months to iterate over the entire dataset. 
-The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution. 
+The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
+of the computations on the nodes of the database cluster. 
 We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/jhtdb_diagram.pdf}
-\caption{Architectrue of the JHTDB.}
-\label{fig:jhtdb_diagram}
-\end{figure}
-
 %Why is it interesting and important?
@@
 capabilities have to be built-in through user-defined functions or stored procedures.
+
 Database systems support rollup queries, including top-$k$ queries, but in most cases these
 queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
-Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning.
+Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
+survey of top-$k$ evaluation strategies).
 This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
@@
 are within the kernel of the computation. 
+Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
+same threshold level will produce different number of points in the result set for different time-steps.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/jhtdb_diagram.pdf}
+\caption{Architectrue of the JHTDB.}
+\label{fig:jhtdb_diagram}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/pdf.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
-\end{figure}
-
 %-The evaluation is computationally limited and embarrassingly parallel, which makes it a good candidate for a GPU implementation.
@@
 \section{JHTDB}\label{jhtdb}
-
-\begin{figure*}[t]
-\centering
-\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
-\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
-\label{fig:max_vort_cluster}
-\end{figure*}
 
@@
 top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
-threshold and top-$k$ queries usually have to examine the entire data volume of a time-step. Furthermore, the data product of threshold and top-$k$ 
+threshold and top-$k$ queries usually have to examine the entire data volume of a time-step or a signification portion of it. Furthermore, the data product 
+of threshold and top-$k$ 
 queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
@@
 at those locations is of the most interest.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/pdf.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
 
@@
 
+\begin{figure*}[t]
+\centering
+\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
+\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
+\label{fig:max_vort_cluster}
+\end{figure*}
+
 Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
@@
 
-{\bf Derived fields computation:} The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
-queries of most interest to science users produce all locations where the values of a derived field are above a given threshold. Thus, the
+{\bf Derived fields computation:} The databases store only the raw field data from the simulation (e.g. velocity, pressure, magnetic field etc.). However,
+the threshold queries of most interest to science users produce all locations where the values of a \emph{derived} field are above a given threshold. Thus, the
 derived field in question has to be computed from the raw data first. For most derived fields of interest, this computation has local support. It has an 
@@
 preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
-and not the raw data fields.
+and not the raw data of the simulation fields.
 Thus, we do not have to derive the requested field from the raw data for 
@@
 an entire $1024^3$ time-step part of a 20TB dataset in less than two minutes. The introduction of an application-aware cache for the query results of these
-queries reduces this time to several seconds when there is a cache hit. In contrast, an evaluation of this functionality performed locally by the user takes
-over 20 hours. To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
+queries reduces this time to several seconds when there is a cache hit. In contrast, one of our science collaborators reported that his evaluation of this
+functionality performed locally would take over 20 hours to complete. 
+To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
 subregions of a time-step. This is necessary as requesting a derived field over an entire time-step will overload the network. Derived fields may have even
@@
 
-\subsection{Execution Time Breakdown}
-\kk{OR}
 \subsection{Evaluation of Additional Fields}
@@
 because its kernel of computation is a single point and therefore there are no additional data along the boundary that have to be requested from adjacent 
-nodes. In that case all of the data needed are available locally.
-
-In all of these cases the time take to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
-time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase with the number of points in
-the result set.
+nodes. In that case all of the data needed are available locally for each database node.
+
+In all of these cases the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
+time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase proportionally to the number
+of points in the result set.
 
@@
 support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
-partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data is 
-handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB is 
-an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping chunks.
-SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named dimensions with
-AQL and make use of the functional operators defined in AFL, such
+partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
+are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
+is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
+chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
+dimensions with AQL and make use of the functional operators defined in AFL, such
 as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
@@
 connection. The Web-services methods can be called from any modern programming language and we provide C, Fortran and Matlab client libraries for the
-JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by mens of a stored procedure
+JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by means of a stored
+procedure
 or a user-defined function that has been implemented and deployed to handle these requests. This allows us to fine-tune the execution of these procedures
@@
 In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which
-will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.
+resides on the servers near the data.
+This will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.
+
+%\section{Acknowledgment}
+
+%The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
+%their insightful comments and suggestions. This work is supported in part by the National Science Foundation
+%under Grants ....
 
@@
 %% %\end{document}  % This is where a 'short' article might terminate
-
-\section{Acknowledgment}
-
-The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
-their insightful comments and suggestions. This work is supported in part by the National Science Foundation
-under Grants ....
 

### 2014-10-07 paper.tex
Bigger fonts for figures. Other minor changes.

@@
 of the computations on the nodes of the database cluster. 
-We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
+We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
 The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
@@
 Database systems support rollup queries, including top-$k$ queries, but in most cases these
-queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
+queries apply only simple linear score functions on the attribute values of individual records.
 Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
@@
 Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
-In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
-are within the kernel of the computation. 
+In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
+are within the kernel of the computation. The functions used to the derive the field may even be non-linear.
 Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
@@
 We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
-stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the array data 
-stored in the database.
+stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the 
+stored array data.
 Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
@@
 
+Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
+locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries and is therefore a 
+limited version of a semantic cache. Nevertheless, it can easily be extended to cache the results of other query types as well if that becomes advantageous.
+
 %\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
@@
 require $\sim$5TB (15TB for vector fields).
+
+The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
+dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
+of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
+the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
+lookup.
+
+{\bf Overall execution of threshold queries:} Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. 
+The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
+First a cache lookup is performed.
+If the data for the requested field, time-step and spatial region
+are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
+retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
+cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested
+from adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is
+compared against the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not
+have enough space for the new records, space is freed up by removing the least recently used data across all quantities. Reading from, updating or
+modifying the cache is done within a transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the 
+cache for each transaction. This provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
 
@@
 
-The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
-dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
-of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
-the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
-lookup.
-
-{\bf Overall execution of threshold queries:} Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
-above the given threshold from the database in the presence of a cache. 
-The mediator submits a query to each of the database nodes storing the raw data
-asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
-First a cache lookup is performed.
-If the data for the requested field, time-step and spatial region
-are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
-retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
-cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested
-from adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is
-compared against the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not
-have enough space for the new records, space is freed up by removing the least recently used data across all quantities. Reading from, updating or
-modifying the cache is done within a transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the 
-cache for each transaction. This provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
-
 \section{Experimental Results}
@@
 \begin{tabular}{|c|c|c|c|c|} \hline
-\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Average Running time (s.)}\\ \cline{3-5}
+\multirow{2}{0.5in}{Vorticity threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Average Running time (s.)}\\ \cline{3-5}
 & & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
@@
 
-As expected we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
-Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
-node with
-the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. SQL Server already makes use of parallelism
-internally and the data have to be retrieved from the same set of disks. Nevertheless, the I/O time does decrease with additional processes, this is because
-the data reside in a partitioned table and the data in each partition are placed in a separate file on one of the disk arrays. Depending on how the data
-requests are scheduled in SQL Server this allows for the disks arrays to be driven in parallel. Additionally, with more processes per node the data can
-also be consumed faster. It is worth noting that the total running time for the queries evaluated with 4 or 8
-processes is about the same as the time it takes to perform the I/O only with a single process.
-
 \begin{figure}
@@
 \end{figure*}
+
+As expected we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
+Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
+node with
+the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. SQL Server already makes use of parallelism
+internally and the data have to be retrieved from the same set of disks. Nevertheless, the I/O time does decrease with additional processes, this is because
+the data reside in a partitioned table and the data in each partition are placed in a separate file on one of the disk arrays. Depending on how the data
+requests are scheduled in SQL Server this allows for the disks arrays to be driven in parallel. Additionally, with more processes per node the data can
+also be consumed faster. It is worth noting that the total running time for the queries evaluated with 4 or 8
+processes is about the same as the time it takes to perform the I/O only with a single process.
 
@@
 idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
-The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
+The caching mechanism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
 at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using

### 2014-10-09 paper.tex
Some changes to bibliography items.
Added reference to "active semantic caching".

@@
 application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
-on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
+on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
 we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
@@
 the cache that we introduce resides on disk rather than in memory, which greatly increases its potential size.
+
+Andrade et al. \cite{Andrade} describe a database system and an optimization framework build on the concept of \emph{active semantic caching}. An
+active semantic cache aims to fully or partially reuse cached query results or aggregates through automated transformations of these aggregates. Similarly
+to our work they focus on real scientific data-analysis applications. The method that we have developed for the evaluation of threshold queries complements
+the active semantic caching approach and could be used in that framework. Our work has focused on extending a relational database system (Microsoft's SQL
+Server) as opposed to designing a new database system from the ground up as is the case in \cite{Andrade}.
 

### 2014-10-20 paper.tex
Better abstract/intro.

@@
 Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
-locally and can take days or months to iterate over the entire dataset. 
+locally by the user taking days or months to iterate over the entire dataset. 
 The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
@@
 The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
+In the case of computational turbulence this allows scientists to quickly focus on the most intense regions in any time-step or the dataset as a whole,
+which greatly speeds up the rate of scientific exploration and discovery. 
+Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
+queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
 \end{abstract}
@@
 fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
-of 1.1PB and 5PB
+of 1.1PB and 11PB
 respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
@@
 Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
-functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-$k$ queries and similarity search in general
-are important in many different disciplines. In this paper we introduce an efficient evaluation strategy for threshold
-queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
+functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
+forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
+in general are important in many different disciplines.We introduce an efficient evaluation strategy for threshold
+queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
 scalar field data stored in the database, but also performs thresholding of derived fields.
@@
 %Why is it hard? (E.g., why do naive approaches fail?)
+Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
+examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
 The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
 The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
-In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
-will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
+In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
+will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
 Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
-but also the transfer of a large amount of data over the network most of which are subsequently discarded. 
+but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
 One of our collaborators reported that such a local evaluation of a threshold
-query over an entire time-step would take over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
-would take months to iterate over the entire dataset. This highlighted the need for providing this important capability through an integrated approach,
+query over an entire time-step took over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
+would take months to iterate over the entire dataset. This highlighted the need for providing the capability through an integrated approach,
 which performs the evaluation server-side.
 
-It is impractical to materialize all possible derived fields with the raw data
+It is impractical to materialize all possible derived fields and store them alongside the raw data
 due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
@@
 
-Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
-shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. This is mainly because these application-independent 
-caches cannot exploit dataset-specific structure and application-level information. Moreover, even if the data are available in one or more of these
-application-independent caches the
-computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.
+Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
+sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
+information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
+with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached.
 
@@
 %Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
-Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
-to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
-datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
-Relational database systems however often do not support
-all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
-capabilities have to be built-in through user-defined functions or stored procedures.
-
 Database systems support rollup queries, including top-$k$ queries, but in most cases these
@@
 \section{JHTDB}\label{jhtdb}
+Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
+to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
+datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
+Relational database systems however often do not support
+all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
+capabilities have to be built-in through user-defined functions or stored procedures.
 
 The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
-Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serve as a public
+Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public
 virtual laboratory for
-the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
+the study of turbulent phenomena. The JHTDB stores several datasets, which are the output of high-resolution numerical simulations of turbulence.
 The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
@@
 differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
-have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
-top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
-threshold and top-$k$ queries usually have to examine the entire data volume of a time-step or a signification portion of it. Furthermore, the data product 
-of threshold and top-$k$ 
-queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
+have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
+of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
+data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
+data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
 previously computed results makes the query results suitable to caching.
@@
 from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. 
-The timespan stored is equal to about one large eddy turnover time. 
+%The timespan stored is equal to about one large eddy turnover time. 
 %It is not clear whether a different more intense event could emerge from an existing
@@
 Figure \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
-There are several worms interacting in a complex way at the same time.
+There are several worms interacting in a complex way at the same time. Similar type of analysis and the fact that the entire time history of the simulation
+is available in a database cluster, which provides built-in sophisticated analysis routines revealed flux-freezing breakdown in MHD turbulence \cite{Eyink},
+showing why solar flares last minutes rather than the millions of years that theory would predict.
 
@@
 locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
-reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries and is therefore a 
-limited version of a semantic cache. Nevertheless, it can easily be extended to cache the results of other query types as well if that becomes advantageous.
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
+easily be extended to cache the results of other query types as well if that becomes advantageous.
 
@@
 
+%\vspace{10 pt}
 \subsection{Experimental Setup}
@@
 
-For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
+For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields that
 have to be computed from the stored data. It is defined as the curl of the velocity field.
@@
 \ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
-the cache first. For these experiments we requested
+the cache first (blue and red bars in the figure). The execution times are also shown in Table \ref{tab:cache_effectiveness}. 
+For these experiments we requested
 the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
@@
 The execution times were averaged over 10 runs.
-We utilized 4 processes per database
-node for the evaluation of each query. Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
-misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
+We utilized 4 processes per database node for the evaluation of each query. 
+The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
+The running time increases slightly only because of the larger result set that has to be returned to the user.
+%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
+%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
 
@@
 
-As expected we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
+As expected, we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
 Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
@@
 
-A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
-queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to
-the time series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time series
-stored in the database (usually one dimensional sequence of measurements) and the time series given in the query. In contrast, our approach focuses on
+A{\ss}falg et al. introduce the concept of threshold queries in time-series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
+queries described in this paper. They are concerned with determining the time-series, which exceed a user-defined threshold at time frames similar to
+the time-series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time-series
+stored in the database (usually one dimensional sequence of measurements) and the time-series given in the query. In contrast, our approach focuses on
 reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.
@@
 the active semantic caching approach and could be used in that framework. Our work has focused on extending a relational database system (Microsoft's SQL
-Server) as opposed to designing a new database system from the ground up as is the case in \cite{Andrade}.
+Server) as opposed to designing a new database system from the ground up as Andrade et al. \cite{Andrade}.
 
@@
 
-In the future we plan to develop
+In the future, we plan to develop
 declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly

### 2014-10-20 paper.tex
Added cache diagram.

@@
 \includegraphics[width=3.35in]{Figures/jhtdb_diagram.pdf}
-\caption{Architectrue of the JHTDB.}
+\caption{Architecture of the JHTDB.}
 \label{fig:jhtdb_diagram}
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/pdf.pdf}
+\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
+\label{fig:pdf}
+\end{figure}
+
 The database nodes are part of the GrayWulf \cite{Szalay} and Data-Scope \cite{DataScope} clusters. Each node
@@
 
-%\begin{figure*}
-%\centering
-%\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
-%\caption{Cache lookup.}
-%\label{fig:cache_diagram}
-%\end{figure*}
-
 The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
@@
 
-\begin{figure}
+\begin{figure*}[t]
 \centering
-\includegraphics[width=3.35in]{Figures/pdf.pdf}
-\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
-\label{fig:pdf}
-\end{figure}
+\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
+\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
+\label{fig:max_vort_cluster}
+\end{figure*}
 
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
-\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
-\label{fig:max_vort_cluster}
-\end{figure*}
-
 Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
@@
 introduction of an application-aware cache
-for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
-answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
-cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
+for query results (Fig. \ref{fig:cache_diagram}). The results of these queries are small compared to the amount of data that need to be examined and the
+results can be used to answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node 
+has a local cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
 used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
 avoid dirty-reads or an inconsistent view of the cache.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/cache_diagram.pdf}
+\caption{Distributed evaluation of threshold queries and architecture of the application-aware cache. 
+Each database node has local cache tables, which reside on solid-state drives attached to the node.}
+\label{fig:cache_diagram}
+\end{figure}
 
@@
 lookup.
-
-{\bf Overall execution of threshold queries:} Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
-above the given threshold from the database in the presence of a cache. 
-The mediator submits a query to each of the database nodes storing the raw data
-asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
-First a cache lookup is performed.
-If the data for the requested field, time-step and spatial region
-are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
-retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
-cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
-parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested
-from adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is
-compared against the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not
-have enough space for the new records, space is freed up by removing the least recently used data across all quantities. Reading from, updating or
-modifying the cache is done within a transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the 
-cache for each transaction. This provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
 
@@
 
+{\bf Overall execution of threshold queries:} Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
+above the given threshold from the database in the presence of a cache. 
+The mediator submits a query to each of the database nodes storing the raw data
+asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. 
+First a cache lookup is performed.
+If the data for the requested field, time-step and spatial region
+are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
+retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
+cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
+parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested
+from adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is
+compared against the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not
+have enough space for the new records, space is freed up by removing the least recently used data across all quantities. Reading from, updating or
+modifying the cache is done within a transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the 
+cache for each transaction. This provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.
+
 \section{Experimental Results}
@@
 
-As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
-less than 3\% and within the margin of error.
-The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
-During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
-produce a cache miss and would have to be evaluated from the raw data. 
-The execution times were averaged over 10 runs.
-We utilized 4 processes per database node for the evaluation of each query. 
-The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
-The running time increases slightly only because of the larger result set that has to be returned to the user.
-%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
-%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
-
 \begin{table}[h]
@@
 \end{table}
+
+As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
+less than 3\% and within the margin of error.
+The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
+During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
+produce a cache miss and would have to be evaluated from the raw data. 
+The execution times were averaged over 10 runs.
+We utilized 4 processes per database node for the evaluation of each query. 
+The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
+The running time increases slightly only because of the larger result set that has to be returned to the user.
+%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
+%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
 
@@
 \ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit}, \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
-SSDs attached to each database node and retrieving the data is always done through a clustered index lookup. In the cases of a cache hit the majority of
-the time is spent simply transferring the results from the database nodes to the mediator and then back to the user. These times remain more or less
-constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
+SSDs attached to each database node (see Fig. \ref{fig:cache_diagram}) and retrieving the data is always done through a clustered index lookup.
+In the cases of a cache hit the majority of the time is spent simply transferring the results from the database nodes to the mediator and then back to the user.
+These times remain more or less constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
 Caching the results of threshold queries
@@
 the active semantic caching approach and could be used in that framework. Our work has focused on extending a relational database system (Microsoft's SQL
-Server) as opposed to designing a new database system from the ground up as Andrade et al. \cite{Andrade}.
+Server) as opposed to designing a new database system from the ground up as described by Andrade et al. \cite{Andrade}.
 

### 2014-10-20 paper.tex
Fixed a typo.

@@
 forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
-in general are important in many different disciplines.We introduce an efficient evaluation strategy for threshold
+in general are important in many different disciplines. We introduce an efficient evaluation strategy for threshold
 queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
-scalar field data stored in the database, but also performs thresholding of derived fields.
+scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.
 
@@
 In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
-are within the kernel of the computation. The functions used to the derive the field may even be non-linear.
+are within the kernel of the computation. The functions used to derive the field may even be non-linear.
 Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the

### 2014-12-29 paper.tex
Addressing reviewers' comments. Added challenges and contributions in introduction. Clarified 10^6 limitation for result sets. Added discussion of sampling approaches in related work section.

@@
 \usepackage{array}
-%\usepackage{natbib}% bold math
 \newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
-%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
 \newcommand{\ddt}[1]{\partial_t #1}
@@
 \definecolor{light-gray}{gray}{0.95}
-
-\lstdefinestyle{cxx}{ %
-language=C++,
-basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
-% numbers=left,                   % where to put the line-numbers
-% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
-% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
-% numbersep=5pt,                  % how far the line-numbers are from the code
-% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
-showspaces=false,               % show spaces adding particular underscores
-commentstyle=\color{cyan},
-stringstyle=\ttfamily\color{red},
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-frame=single,                 % adds a frame around the code
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to bottom
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
-% morekeywords={char,double},
-escapeinside={\%*}{*)},          % if you want to add a comment within your code
-% morekeywords={*,...}          % if you want to add more keywords to the set
-% morekeywords={char,double}    % if you want to add more keywords to the set
-frameround=fttt
-}
-
-\lstdefinestyle{fortran}{ %
-language=Fortran,
-basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
-% numbers=left,                   % where to put the line-numbers
-% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
-% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
-% numbersep=5pt,                  % how far the line-numbers are from the code
-% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
-showspaces=false,               % show spaces adding particular underscores
-commentstyle=\color{cyan},
-stringstyle=\ttfamily\color{red},
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-frame=single,                 % adds a frame around the code
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to bottom
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
-% morekeywords={char,double},
-escapeinside={\%*}{*)},          % if you want to add a comment within your code
-% morekeywords={*,...}          % if you want to add more keywords to the set
-% morekeywords={char,double}    % if you want to add more keywords to the set
-frameround=fttt
-}
 
@@
 
-%\conferenceinfo{EuroMPI}{2013 Madrid, Spain}
-
 \title{
@@
 }
-%
-% You need the command \numberofauthors to handle the 'placement
-% and alignment' of the authors beneath the title.
-%
-% For aesthetic reasons, we recommend 'three authors at a time'
-% i.e. three 'name/affiliation blocks' be placed beneath the title.
-%
-% NOTE: You are NOT restricted in how many 'rows' of
-% "name/affiliations" may appear. We just ask that you restrict
-% the number of 'columns' to three.
-%
-% Because of the available 'opening page real-estate'
-% we ask you to refrain from putting more than six authors
-% (two rows with three columns) beneath the article title.
-% More than six makes the first-page appear very cluttered indeed.
-%
-% Use the \alignauthor commands to handle the names
-% and affiliations for an 'aesthetic maximum' of six authors.
-% Add names, affiliations, addresses for
-% the seventh etc. author(s) as the argument for the
-% \additionalauthors command.
-% These 'additional authors' will be output/set for you
-% without further effort on your part as the last section in
-% the body of your article BEFORE References or any Appendices.
 
-\numberofauthors{3} %  in this sample file, there are a *total*
-% of EIGHT authors. SIX appear on the 'first-page' (for formatting
-% reasons) and the remaining two appear in the \additionalauthors section.
-%
+\numberofauthors{3} 
 \author{
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
 \alignauthor Kalin Kanov \\
@@
 }
-%% \and % use '\and' if you need 'another row' of author names
-%% % There's nothing stopping you putting the seventh, eighth, etc.
-%% % author on the opening page (as the 'third row') but we ask,
-%% % for aesthetic reasons that you place these 'additional authors'
-%% % in the \additional authors block, viz.
-%% \additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
-%% email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
-%% (The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
-%% \date{30 July 1999}
-%% % Just remember to make sure that the TOTAL number of authors
-%% % is the number that will appear on the first page PLUS the
-%% % number that will appear in the \additionalauthors section.
 
@@
 The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
-of the computations on the nodes of the database cluster. 
-We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
+of the computations on the nodes of an analysis database cluster. 
+We extend the scientific analysis environment with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
 The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
-In the case of computational turbulence this allows scientists to quickly focus on the most intense regions in any time-step or the dataset as a whole,
-which greatly speeds up the rate of scientific exploration and discovery. 
+In the case of computational turbulence this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
+the dataset as a whole, which greatly speeds up the rate of scientific exploration and discovery. 
 Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
@@
 %% % A category with the (minimum) three required fields
-%\category{}{Extending Data Base Technology}
+\category{H.2.8}{Database Management}{Database Applications -- scientific, spatial databases and GIS}
 %% %A category including the fourth, optional field follows...
@@
 to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
-fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
-of 1.1PB and 11PB
+fast access to such data sets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
+Johns Hopkins, which have capacity of 1.1PB and 11PB
 respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
-These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
+These systems differ from the traditional HPC environments in that they aim to achieve high aggregate throughput by balancing computation
 capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
@@
 queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
-scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.
+scalar field data stored in the database, but also performs thresholding of \emph{derived} fields. The main challenge that our approach tackles is that the
+data-intensive computation of \emph{derived} fields has to be carried out on-demand for extremely large array datasets stored in an analysis cluster environment 
+comprised of multiple database nodes. Here we focus on the evaluation of threshold queries over the fields derived from the data stored in the cluster as
+these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
+queries in general.
 
@@
 examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
-The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
+The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
 The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
@@
 %Why is it interesting and important?
-Thresholding allows scientists to obtain and examine the most intense regions and features in the dataset in the case of turbulence. 
+Thresholding allows scientists to obtain and examine the regions containing the most intense events and features in the dataset in the case of turbulence. 
 These are often the locations that have the largest vorticity norms and have intense vortices or reconnection events. 
@@
 %Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
-Database systems support rollup queries, including top-$k$ queries, but in most cases these
-queries apply only simple linear score functions on the attribute values of individual records.
-Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
-survey of top-$k$ evaluation strategies).
-This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
-Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
-In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
-are within the kernel of the computation. The functions used to derive the field may even be non-linear.
-Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
-same threshold level will produce different number of points in the result set for different time-steps.
+There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
+evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
+large multidimensional array datasets. Such a stencil kernel computation has to be performed at each location on the grid as opposed to at a particular number 
+of target locations. Finally, the evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data across 
+the network and to achieve scalability.
 
@@
 We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
-stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the 
+stored in the database. The derived fields, which our method computes and operates on are derived by performing stencil kernel computations over the 
 stored array data.
@@
 The cache that we describe is therefore a limited version of a semantic cache.
+
+The contributions of this paper are the following:
+\begin{itemize}
+\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. For the first time providing large
+data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
+\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching and taking advantage of heterogeneous architectures (sharded database cluster with several SSDs per node).
+\item Evaluating the proposed method on data-intensive workloads in a live production environment and showing scalability results on datasets hundreds of terabytes in size.
+\end{itemize}
 
@@
 The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
-Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public
-virtual laboratory for
+Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. It solves a pressing problem in a production scientific analysis environment, which differs 
+from the traditional supercomputing environments and provides large data analytics capabilities transparently to the public. 
+The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public virtual laboratory for
 the study of turbulent phenomena. The JHTDB stores several datasets, which are the output of high-resolution numerical simulations of turbulence.
@@
 
-The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
+The JHTDB provides a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
 differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
@@
 against the specified threshold and if the value is higher it is maintained along with
-the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
-the entire time-step or a significant fraction of it for queries with thresholds that are set too low.
+the spatial coordinates of the location in a list.
 
@@
 
-Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
-locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
-reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
-easily be extended to cache the results of other query types as well if that becomes advantageous.
-
-%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
-%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
-%computation accessing multiple data points in a kernel around the target location.}
-
-The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
-Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
-space and database overhead is $\sim$40MB. 
-Therefore for a dataset containing 
-1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
-require $\sim$5TB (15TB for vector fields).
-
-The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
-dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
-of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
-the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
-lookup.
-
 \begin{algorithm}
@@
 \State $ordinal \gets reader[``ordinal"]$
-\If {$k \le k_s \And q \in \text{[start, end]}$}
+\If {$k \ge k_s \And q \in \text{[start, end]}$}
 \State $\textit{query} \gets \text{SELECT * FROM cachedb..cacheData}$
@@
 \end{algorithm}
+
+Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
+locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
+easily be extended to cache the results of other query types as well if that becomes advantageous.
+
+We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around 1,000,000 locations, which is
+sufficient to examine a time-step in detail while limiting the amount of data that have to be returned to the user over the network as well as the amount of
+data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
+is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
+of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.
+
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
+space and database overhead is $\sim$40MB. 
+Therefore for a dataset containing 
+1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+require $\sim$5TB (15TB for vector fields).
+
+The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
+dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
+of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
+the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
+lookup.
 
@@
 thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
-the development of the most intense vortices that are observed. 
-%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}
-
-\subsection{Evaluation of cache effectiveness}
+the development of the most intense vortices observed in the dataset.
 
@@
 
+\subsection{Evaluation of cache effectiveness}
+
 The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
@@
 (or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
-(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
+(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set and there were $\sim$900,000 points 
 (or 0.0847\% of all points) above the threshold. For each set a random
@@
 Cache hits reduce the running time of threshold queries by over an order of magnitude as shown in Figure \ref{fig:cache_overhead} and
-Table \ref{tab:cache_effectiveness}. This is because we do not have to compute the derived
-field requested from the raw data, which also eliminates the associated I/O. Only the cache entries need to be looked-up, which is substantially less data
+Table \ref{tab:cache_effectiveness}. This is because we do not have to compute the requested derived
+field from the raw data, which eliminates the associated I/O. Only the cache entries need to be looked-up, which is substantially less data
 than the raw vector or scalar field data. For the queries with large result sets it is actually the network time taken to transfer the results to the user that 
@@
 measure their running times. Let us focus on the query
-with low threshold, which returns $\sim$900,000 points. Given that the number of points that can be returned from a threshold query is limited to 1,000,000
+with low threshold, which returns $\sim$900,000 points. Given that valid threshold values are limited to those that result in no more than 1,000,000 points
 it is likely that all subsequent queries to this time-step will result in a cache hit as their threshold is likely to be equal or higher than the cached one.
@@
 the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
-compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
+compute the derived field from data available locally with only a small amount along the boundary of each region having to be retrieved from adjacent nodes.
 Each computation is independent and embarrassingly parallel. This allows us to make use of multiple processes per node and scale out to
@@
 Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
-the boundary of each region will be requested by multiple processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
-compared to running a single process.
+the boundary of each region are requested by multiple processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
+compared to running with a single process.
 
@@
 The vorticity field and the Q-criterion have similar I/O requirements as they
-have the same kernels of computation and both can be derived from the velocity gradient. The vorticity has 3 components and its computation only examines
+have the same kernels of computation and are both derived from the velocity gradient. The vorticity has 3 components and its computation only examines
 6 of the 9 components of the velocity gradient, which are also examined in pairs (see Eq. \ref{eq:curl}). On the other hand, even though the Q-criterion is a
@@
 
+Database systems support rollup queries, including top-$k$ queries, but in most cases these
+queries apply only simple linear score functions on the attribute values of individual records.
+Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
+survey of top-$k$ evaluation strategies).
+This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
+Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
+In contrast our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
+examines the vector or scalar array data at all neighboring locations, which
+are within the kernel of the computation. The functions used to derive the field may even be non-linear.
+Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
+same threshold level will produce different number of points in the result set for different time-steps.
+
 The processing of top-$k$ queries has been studied extensively in the context of distributed and relational database systems. A survey of different techniques
@@
 \cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deals with array data stored in
-a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the attribute values of
-individual records. In addition to top-$k$ queries, we are also interested in threshold queries, where the result set includes all locations and values that are
-above a user-prescribed threshold. The threshold queries that we focus on operate over fields derived from the scalar or vector field simulation data through
-the evaluation of kernel computations at each data point.
-
-Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
-idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
-The caching mechanism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
-at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
-simple linear score functions.
+a relational database system. Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA
+\cite{Zhao}. They build on the idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect 
+to the data stored on each server. The caching mechanism that we use is similar in that regard, but the queries that are evaluated in our system operate on
+derived fields, which are computed at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the 
+attributes of individual records only using simple linear score functions.
 
@@
 perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache,
-we do not cache raw data objects, but rather query results. This substantially reduces the size of the cache and improves query performance. Additionally, 
-the cache that we introduce resides on disk rather than in memory, which greatly increases its potential size.
+we do not cache raw data objects, but rather query results. Caching query results preserves the computational effort in addition to reducing I/O, which has a
+much bigger impact on query performance and substantially reduces the size of the cache.. Additionally, the cache that we introduce resides on disk rather 
+than in memory, which greatly increases its potential size. Lopez et al. also explore approximate querying through aggregation, which can be fairly easily
+supported by our system but is of limited use as scientists performing threshold queries are usually interested in obtaining the exact locations where a field is 
+at its highest values.
+
+Sampling approaches \cite{Sidirourgos, Nirkhiwale} offer an alternative to the computation of derived fields on demand and the evaluation of threshold queries
+on them. The goal of both techniques is to not return large data volumes, but focus on the most intense events and interesting regions in the dataset. 
+The computation of the derived fields is carried out on the nodes of the database cluster and takes a look at the dataset as a whole, while the user obtains only
+a small subset of the data, where the derived field in question is above the prescribed threshold. Sampling approaches can potentially omit some locations and
+while useful for generating initial impressions may not be suitable if the exact locations where a field is at its highest values are desired.
 
@@
 
-%% \section{The {\secit Body} of The Paper}
-%% Typically, the body of a paper is organized
-%% into a hierarchical structure, with numbered or unnumbered
-%% headings for sections, subsections, sub-subsections, and even
-%% smaller sections.  The command \texttt{{\char'134}section} that
-%% precedes this paragraph is part of such a
-%% hierarchy.\footnote{This is the second footnote.  It
-%% starts a series of three footnotes that add nothing
-%% informational, but just give an idea of how footnotes work
-%% and look. It is a wordy one, just so you see
-%% how a longish one plays out.} \LaTeX\ handles the numbering
-%% and placement of these headings for you, when you use
-%% the appropriate heading commands around the titles
-%% of the headings.  If you want a sub-subsection or
-%% smaller part to be unnumbered in your output, simply append an
-%% asterisk to the command name.  Examples of both
-%% numbered and unnumbered headings will appear throughout the
-%% balance of this sample document.
+\bibliographystyle{abbrv}
+\bibliography{paper}
 
-%% Because the entire article is contained in
-%% the \textbf{document} environment, you can indicate the
-%% start of a new paragraph with a blank line in your
-%% input file; that is why this sentence forms a separate paragraph.
-
-%% \subsection{Type Changes and {\subsecit Special} Characters}
-%% We have already seen several typeface changes in this sample.  You
-%% can indicate italicized words or phrases in your text with
-%% the command \texttt{{\char'134}textit}; emboldening with the
-%% command \texttt{{\char'134}textbf}
-%% and typewriter-style (for instance, for computer code) with
-%% \texttt{{\char'134}texttt}.  But remember, you do not
-%% have to indicate typestyle changes when such changes are
-%% part of the \textit{structural} elements of your
-%% article; for instance, the heading of this subsection will
-%% be in a sans serif\footnote{A third footnote, here.
-%% Let's make this a rather short one to
-%% see how it looks.} typeface, but that is handled by the
-%% document class file. Take care with the use
-%% of\footnote{A fourth, and last, footnote.}
-%% the curly braces in typeface changes; they mark
-%% the beginning and end of
-%% the text that is to be in the different typeface.
-
-%% You can use whatever symbols, accented characters, or
-%% non-English characters you need anywhere in your document;
-%% you can find a complete list of what is
-%% available in the \textit{\LaTeX\
-%% User's Guide}\cite{Lamport:LaTeX}.
-
-%% \subsection{Math Equations}
-%% You may want to display math equations in three distinct styles:
-%% inline, numbered or non-numbered display.  Each of
-%% the three are discussed in the next sections.
-
-%% \subsubsection{Inline (In-text) Equations}
-%% A formula that appears in the running text is called an
-%% inline or in-text formula.  It is produced by the
-%% \textbf{math} environment, which can be
-%% invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
-%% construction or with the short form \texttt{\$. . .\$}. You
-%% can use any of the symbols and structures,
-%% from $\alpha$ to $\omega$, available in
-%% \LaTeX\cite{Lamport:LaTeX}; this section will simply show a
-%% few examples of in-text equations in context. Notice how
-%% this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
-%% set here in in-line math style, looks slightly different when
-%% set in display style.  (See next section).
-
-%% \subsubsection{Display Equations}
-%% A numbered display equation -- one set off by vertical space
-%% from the text and centered horizontally -- is produced
-%% by the \textbf{equation} environment. An unnumbered display
-%% equation is produced by the \textbf{displaymath} environment.
-
-%% Again, in either environment, you can use any of the symbols
-%% and structures available in \LaTeX; this section will just
-%% give a couple of examples of display equations in context.
-%% First, consider the equation, shown as an inline equation above:
-%% \begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
-%% Notice how it is formatted somewhat differently in
-%% the \textbf{displaymath}
-%% environment.  Now, we'll enter an unnumbered equation:
-%% \begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
-%% and follow it with another numbered equation:
-%% \begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
-%% just to demonstrate \LaTeX's able handling of numbering.
-
-%% \subsection{Citations}
-%% Citations to articles \cite{bowman:reasoning, clark:pct, braams:babel, herlihy:methodology},
-%% conference
-%% proceedings \cite{clark:pct} or books \cite{salas:calculus, Lamport:LaTeX} listed
-%% in the Bibliography section of your
-%% article will occur throughout the text of your article.
-%% You should use BibTeX to automatically produce this bibliography;
-%% you simply need to insert one of several citation commands with
-%% a key of the item cited in the proper location in
-%% the \texttt{.tex} file \cite{Lamport:LaTeX}.
-%% The key is a short reference you invent to uniquely
-%% identify each work; in this sample document, the key is
-%% the first author's surname and a
-%% word from the title.  This identifying key is included
-%% with each item in the \texttt{.bib} file for your article.
-
-%% The details of the construction of the \texttt{.bib} file
-%% are beyond the scope of this sample document, but more
-%% information can be found in the \textit{Author's Guide},
-%% and exhaustive details in the \textit{\LaTeX\ User's
-%% Guide}\cite{Lamport:LaTeX}.
-
-%% This article shows only the plainest form
-%% of the citation command, using \texttt{{\char'134}cite}.
-%% This is what is stipulated in the SIGS style specifications.
-%% No other citation format is endorsed.
-
-%% \subsection{Tables}
-%% Because tables cannot be split across pages, the best
-%% placement for them is typically the top of the page
-%% nearest their initial cite.  To
-%% ensure this proper ``floating'' placement of tables, use the
-%% environment \textbf{table} to enclose the table's contents and
-%% the table caption.  The contents of the table itself must go
-%% in the \textbf{tabular} environment, to
-%% be aligned properly in rows and columns, with the desired
-%% horizontal and vertical rules.  Again, detailed instructions
-%% on \textbf{tabular} material
-%% is found in the \textit{\LaTeX\ User's Guide}.
-
-%% Immediately following this sentence is the point at which
-%% Table 1 is included in the input file; compare the
-%% placement of the table here with the table in the printed
-%% dvi output of this document.
-
-%% \begin{table}
-%% \centering
-%% \caption{Frequency of Special Characters}
-%% \begin{tabular}{|c|c|l|} \hline
-%% Non-English or Math&Frequency&Comments\\ \hline
-%% \O & 1 in 1,000& For Swedish names\\ \hline
-%% $\pi$ & 1 in 5& Common in math\\ \hline
-%% \$ & 4 in 5 & Used in business\\ \hline
-%% $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-%% \hline\end{tabular}
-%% \end{table}
-
-%% To set a wider table, which takes up the whole width of
-%% the page's live area, use the environment
-%% \textbf{table*} to enclose the table's contents and
-%% the table caption.  As with a single-column table, this wide
-%% table will ``float" to a location deemed more desirable.
-%% Immediately following this sentence is the point at which
-%% Table 2 is included in the input file; again, it is
-%% instructive to compare the placement of the
-%% table here with the table in the printed dvi
-%% output of this document.
-
-%% \begin{table*}
-%% \centering
-%% \caption{Some Typical Commands}
-%% \begin{tabular}{|c|c|l|} \hline
-%% Command&A Number&Comments\\ \hline
-%% \texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
-%% \texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
-%% \texttt{{\char'134}table}& 300 & For tables\\ \hline
-%% \texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
-%% \end{table*}
-%% % end the environment with {table*}, NOTE not {table}!
-
-%% \subsection{Figures}
-%% Like tables, figures cannot be split across pages; the
-%% best placement for them
-%% is typically the top or the bottom of the page nearest
-%% their initial cite.  To ensure this proper ``floating'' placement
-%% of figures, use the environment
-%% \textbf{figure} to enclose the figure and its caption.
-
-%% This sample document contains examples of \textbf{.eps}
-%% and \textbf{.ps} files to be displayable with \LaTeX.  More
-%% details on each of these is found in the \textit{Author's Guide}.
-
-%% \begin{figure}
-%% \centering
-%% \epsfig{file=fly.eps}
-%% \caption{A sample black and white graphic (.eps format).}
-%% \end{figure}
-
-%% \begin{figure}
-%% \centering
-%% \epsfig{file=fly.eps, height=1in, width=1in}
-%% \caption{A sample black and white graphic (.eps format)
-%% that has been resized with the \texttt{epsfig} command.}
-%% \end{figure}
-
-%% As was the case with tables, you may want a figure
-%% that spans two columns.  To do this, and still to
-%% ensure proper ``floating'' placement of tables, use the environment
-%% \textbf{figure*} to enclose the figure and its caption.
-
-%% Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
-%% used; use
-%% the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
-%% commands as appropriate for the different file types.
-
-%% \subsection{Theorem-like Constructs}
-%% Other common constructs that may occur in your article are
-%% the forms for logical constructs like theorems, axioms,
-%% corollaries and proofs.  There are
-%% two forms, one produced by the
-%% command \texttt{{\char'134}newtheorem} and the
-%% other by the command \texttt{{\char'134}newdef}; perhaps
-%% the clearest and easiest way to distinguish them is
-%% to compare the two in the output of this sample document:
-
-%% This uses the \textbf{theorem} environment, created by
-%% the\linebreak\texttt{{\char'134}newtheorem} command:
-%% \newtheorem{theorem}{Theorem}
-%% \begin{theorem}
-%% Let $f$ be continuous on $[a,b]$.  If $G$ is
-%% an antiderivative for $f$ on $[a,b]$, then
-%% \begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
-%% \end{theorem}
-
-%% The other uses the \textbf{definition} environment, created
-%% by the \texttt{{\char'134}newdef} command:
-%% \newdef{definition}{Definition}
-%% \begin{definition}
-%% If $z$ is irrational, then by $e^z$ we mean the
-%% unique number which has
-%% logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
-%% \end{definition}
-
-%% \begin{figure}
-%% \centering
-%% \psfig{file=rosette.ps, height=1in, width=1in,}
-%% \caption{A sample black and white graphic (.ps format) that has
-%% been resized with the \texttt{psfig} command.}
-%% \end{figure}
-
-%% Two lists of constructs that use one of these
-%% forms is given in the
-%% \textit{Author's  Guidelines}.
-
-%% \begin{figure*}
-%% \centering
-%% \epsfig{file=flies.eps}
-%% \caption{A sample black and white graphic (.eps format)
-%% that needs to span two columns of text.}
-%% \end{figure*}
-%% and don't forget to end the environment with
-%% {figure*}, not {figure}!
-
-%% There is one other similar construct environment, which is
-%% already set up
-%% for you; i.e. you must \textit{not} use
-%% a \texttt{{\char'134}newdef} command to
-%% create it: the \textbf{proof} environment.  Here
-%% is a example of its use:
-%% \begin{proof}
-%% Suppose on the contrary there exists a real number $L$ such that
-%% \begin{displaymath}
-%% \lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
-%% \end{displaymath}
-%% Then
-%% \begin{displaymath}
-%% l=\lim_{x\rightarrow c} f(x)
-%% = \lim_{x\rightarrow c}
-%% \left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
-%% = \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
-%% \frac{f(x)}{g(x)} = 0\cdot L = 0,
-%% \end{displaymath}
-%% which contradicts our assumption that $l\neq 0$.
-%% \end{proof}
-
-%% Complete rules about using these environments and using the
-%% two different creation commands are in the
-%% \textit{Author's Guide}; please consult it for more
-%% detailed instructions.  If you need to use another construct,
-%% not listed therein, which you want to have the same
-%% formatting as the Theorem
-%% or the Definition\cite{salas:calculus} shown above,
-%% use the \texttt{{\char'134}newtheorem} or the
-%% \texttt{{\char'134}newdef} command,
-%% respectively, to create it.
-
-%% \subsection*{A {\secit Caveat} for the \TeX\ Expert}
-%% Because you have just been given permission to
-%% use the \texttt{{\char'134}newdef} command to create a
-%% new form, you might think you can
-%% use \TeX's \texttt{{\char'134}def} to create a
-%% new command: \textit{Please refrain from doing this!}
-%% Remember that your \LaTeX\ source code is primarily intended
-%% to create camera-ready copy, but may be converted
-%% to other forms -- e.g. HTML. If you inadvertently omit
-%% some or all of the \texttt{{\char'134}def}s recompilation will
-%% be, to say the least, problematic.
-
-%% \section{Conclusions}
-%% This paragraph will end the body of this sample document.
-%% Remember that you might still have Acknowledgments or
-%% Appendices; brief samples of these
-%% follow.  There is still the Bibliography to deal with; and
-%% we will make a disclaimer about that here: with the exception
-%% of the reference to the \LaTeX\ book, the citations in
-%% this paper are to articles which have nothing to
-%% do with the present subject and are used as
-%% examples only.
-%% %\end{document}  % This is where a 'short' article might terminate
-
-%% %ACKNOWLEDGMENTS are optional
-%% \section{Acknowledgments}
-%% This section is optional; it is a location for you
-%% to acknowledge grants, funding, editing assistance and
-%% what have you.  In the present case, for example, the
-%% authors would like to thank Gerald Murray of ACM for
-%% his help in codifying this \textit{Author's Guide}
-%% and the \textbf{.cls} and \textbf{.tex} files that it describes.
-
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
-\bibliographystyle{abbrv}
-\bibliography{paper}  % sigproc.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
-%
-% ACM needs 'a single self-contained file'!
-%
-%APPENDICES are optional
-%\balancecolumns
-%% \appendix
-%% %Appendix A
-%% \section{Headings in Appendices}
-%% The rules about hierarchical headings discussed above for
-%% the body of the article are different in the appendices.
-%% In the \textbf{appendix} environment, the command
-%% \textbf{section} is used to
-%% indicate the start of each Appendix, with alphabetic order
-%% designation (i.e. the first is A, the second B, etc.) and
-%% a title (if you include one).  So, if you need
-%% hierarchical structure
-%% \textit{within} an Appendix, start with \textbf{subsection} as the
-%% highest level. Here is an outline of the body of this
-%% document in Appendix-appropriate form:
-%% \subsection{Introduction}
-%% \subsection{The Body of the Paper}
-%% \subsubsection{Type Changes and  Special Characters}
-%% \subsubsection{Math Equations}
-%% \paragraph{Inline (In-text) Equations}
-%% \paragraph{Display Equations}
-%% \subsubsection{Citations}
-%% \subsubsection{Tables}
-%% \subsubsection{Figures}
-%% \subsubsection{Theorem-like Constructs}
-%% \subsubsection*{A Caveat for the \TeX\ Expert}
-%% \subsection{Conclusions}
-%% \subsection{Acknowledgments}
-%% \subsection{Additional Authors}
-%% This section is inserted by \LaTeX; you do not insert it.
-%% You just add the names and information in the
-%% \texttt{{\char'134}additionalauthors} command at the start
-%% of the document.
-%% \subsection{References}
-%% Generated by bibtex from your ~.bib file.  Run latex,
-%% then bibtex, then latex twice (to resolve references)
-%% to create the ~.bbl file.  Insert that ~.bbl file into
-%% the .tex source file and comment out
-%% the command \texttt{{\char'134}thebibliography}.
-%% % This next section command marks the start of
-%% % Appendix B, and does not continue the present hierarchy
-%% \section{More Help for the Hardy}
-%% The acm\_proc\_article-sp document class file itself is chock-full of succinct
-%% and helpful comments.  If you consider yourself a moderately
-%% experienced to expert user of \LaTeX, you may find reading
-%% it useful but please remember not to change it.
-%% \balancecolumns
-%% % That's all folks!
 \end{document}

### 2015-01-04 paper.tex
More work on the introduction and justification for the 10^6 limitation for query results.

@@
 Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
-to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
-fast access to such data sets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
+to manage increasingly large datasets. Data-intensive systems and architectures have been developed with the goal of storing and providing
+fast access to such datasets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
 Johns Hopkins, which have capacity of 1.1PB and 11PB
@@
 strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
-databases and in order to reduce their running times they have to make efficient use of the cluster resources.
+databases. In order to reduce their running times they have to both make efficient use of the cluster resources and incorporate 
+leading data management techniques.
 
@@
 queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
-scalar field data stored in the database, but also performs thresholding of \emph{derived} fields. The main challenge that our approach tackles is that the
+scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.
+
+The main challenge that our approach tackles is that the
 data-intensive computation of \emph{derived} fields has to be carried out on-demand for extremely large array datasets stored in an analysis cluster environment 
-comprised of multiple database nodes. Here we focus on the evaluation of threshold queries over the fields derived from the data stored in the cluster as
+comprised of multiple database nodes. We focus on the evaluation of threshold queries of fields derived from the data stored in the cluster as
 these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
-queries in general.
+queries in general. The established data management techniques that our approach combines make it easy to understand. It can be applied to other 
+scientific analysis environments, which manage large datasets in a database management system. Examples include the Sloan Digital Sky Survey
+\cite{SDSS}, the Millennium Simulation \cite{Millennium} and the Open Connectome Project \cite{Burns}.
 
@@
 examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
-The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
-The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
-In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
-will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
+%The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
+%The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
+%In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
+%will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
+It is impractical to materialize all possible derived fields and store them alongside the raw data
+due to the large size of the datasets and the limits of available storage.
 Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
@@
 One of our collaborators reported that such a local evaluation of a threshold
-query over an entire time-step took over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
-would take months to iterate over the entire dataset. This highlighted the need for providing the capability through an integrated approach,
-which performs the evaluation server-side.
+query over an entire time-step took over 20 hours. It would take months to iterate over the entire dataset. 
+This highlighted the need for providing the capability through an integrated approach, which performs the evaluation server-side.
 
-It is impractical to materialize all possible derived fields and store them alongside the raw data
-due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
-have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
-We will demonstrate that this integrated approach evaluates a threshold query over an entire time-step in a few minutes.
+%It is impractical to materialize all possible derived fields and store them alongside the raw data
+%due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
+%have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
+%We will demonstrate that this integrated approach evaluates a threshold query over an entire time-step in a few minutes.
 
@@
 with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached. 
+We will demonstrate that an integrated approach, which computes the derived fields on-demand in a data-parallel manner, performs the evaluation over 
+an entire time-step in a few minutes. Storing the query results in an application-aware semantic cache further reduces the running times to several 
+seconds.
 
@@
 evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
-large multidimensional array datasets. Such a stencil kernel computation has to be performed at each location on the grid as opposed to at a particular number 
-of target locations. Finally, the evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data across 
+large multidimensional array datasets. Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number 
+of target locations. 
+The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
 the network and to achieve scalability.
+Techniques that target the traditional supercomputing environments do not translate directly to the 
+distributed database setting of an analysis cluster environment.
 
@@
 We exploit the parallelism 
-available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
-is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
-system with an
+available in the cluster by means of data-parallel execution of the computations. 
+%Each computation is distributed across the nodes of the cluster and
+%is executed in parallel by several processes within a database node by creating multiple connections to each node. 
+We extend the database management system with an
 application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
 on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
-we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
-The cache that we describe is therefore a limited version of a semantic cache.
+we cache the actual results of the computation in a limited version of a semantic cache. 
+This leads to query performance improvement of over an order of magnitude.
 
@@
 data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
-\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching and taking advantage of heterogeneous architectures (sharded database cluster with several SSDs per node).
+\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching as well as taking
+advantage of heterogeneous scientific cluster architectures (sharded relational DBMS with several SSDs per node).
 \item Evaluating the proposed method on data-intensive workloads in a live production environment and showing scalability results on datasets hundreds of terabytes in size.
@@
 
+Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
+locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
+easily be extended to cache the results of other query types as well if that becomes advantageous.
+
 \begin{algorithm}
@@
 
-Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
-locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
-reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
-easily be extended to cache the results of other query types as well if that becomes advantageous.
-
 We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
-the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around 1,000,000 locations, which is
-sufficient to examine a time-step in detail while limiting the amount of data that have to be returned to the user over the network as well as the amount of
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around $10^6$ locations, which is
+sufficient to examine a time-step in detail. In the case of the vorticity these locations contain all values within $\sim$8 times the root mean squared value of
+the vorticity, which is about $25\%$ of the maximum. Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ 
+of the maximum would be sufficient. 
+At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
 data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
@@
 The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
-Given a limit of $\sim$1,000,000 points per time-step for a threshold query, the space required to cache the maximum number of points including the index
+Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
 space and database overhead is $\sim$40MB. 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
+\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
+absence of a cache.}
+\label{fig:cache_overhead}
+\end{figure}
+
 \section{Experimental Results}
@@
 database file.
-
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
-absence of a cache.}
-\label{fig:cache_overhead}
-\end{figure}
 
@@
 The computations for all of 
-the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
+the derived fields of interest (such as the vorticity) at each grid point need data from adjacent grid points only. Therefore, each node of the cluster is able to
 compute the derived field from data available locally with only a small amount along the boundary of each region having to be retrieved from adjacent nodes.
@@
 
-We observer nearly a two times speedup when going from a single process per node to two processes per node (Fig. \ref{fig:scaleup}). 
+We observe nearly a two times speedup when going from a single process per node to two processes per node (Fig. \ref{fig:scaleup}). 
 The speedup diminishes to 1.4 times when going to 4 processes and little speedup is 
@@
 scalar value, it is computed through a non-linear combination of all 9 of the components of the velocity gradient. This means that the velocity gradient has
-to be computed at each grid location before the Q-criterion is evaluated, which is reflected in the increased computation time that we observer for the
+to be computed at each grid location before the Q-criterion is evaluated, which is reflected in the increased computation time that we observe for the
 Q-criterion. The magnetic field is one of the raw fields of the magnetohydrodynamics dataset that are stored in the database. Therefore, there is no
@@
 
-Sampling approaches \cite{Sidirourgos, Nirkhiwale} offer an alternative to the computation of derived fields on demand and the evaluation of threshold queries
+Sampling approaches \cite{Sidirourgos, Nirkhiwale} offer an alternative to the on-demand computation of derived fields and the evaluation of threshold queries
 on them. The goal of both techniques is to not return large data volumes, but focus on the most intense events and interesting regions in the dataset. 
@@
 
-%\section{Acknowledgment}
-
-%The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
-%their insightful comments and suggestions. This work is supported in part by the National Science Foundation
-%under Grants ....
+\section{Acknowledgment}
+The authors would like to thank the Turbulence Database Group at Johns Hopkins University as well as three anonymous reviewers for
+their insightful comments and suggestions. This work is supported in part by the National Science Foundation
+under Grants ....
 

### 2015-01-05 paper.tex
Fixed some typos.

@@
 
-{\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 
-introduction of an application-aware cache
-for query results (Fig. \ref{fig:cache_diagram}). The results of these queries are small compared to the amount of data that need to be examined and the
-results can be used to answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node 
-has a local cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
-used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
-avoid dirty-reads or an inconsistent view of the cache.
+We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around $10^6$ locations, which is
+sufficient to examine a time-step in detail. In the case of the vorticity these locations contain all values above $\sim$8 times the root mean square value of
+the vorticity, which is about $25\%$ of the maximum. Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ 
+of the maximum would be sufficient. 
+At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
+data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
+is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
+of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.
 
@@
 
+{\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 
+introduction of an application-aware cache
+for query results (Fig. \ref{fig:cache_diagram}). The results of these queries are small compared to the amount of data that need to be examined and the
+results can be used to answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node 
+has a local cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
+used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
+avoid dirty-reads or an inconsistent view of the cache.
+
 Caching the query results
@@
 a small set of data and do not need to perform any additional computation.
-
-Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
-locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
-reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
-easily be extended to cache the results of other query types as well if that becomes advantageous.
 
@@
 
-We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
-the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around $10^6$ locations, which is
-sufficient to examine a time-step in detail. In the case of the vorticity these locations contain all values within $\sim$8 times the root mean squared value of
-the vorticity, which is about $25\%$ of the maximum. Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ 
-of the maximum would be sufficient. 
-At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
-data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
-is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
-of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.
+Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
+locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
+easily be extended to cache the results of other query types as well if that becomes advantageous.
 
@@
 we do not cache raw data objects, but rather query results. Caching query results preserves the computational effort in addition to reducing I/O, which has a
-much bigger impact on query performance and substantially reduces the size of the cache.. Additionally, the cache that we introduce resides on disk rather 
+much bigger impact on query performance and substantially reduces the size of the cache. Additionally, the cache that we introduce resides on disk rather 
 than in memory, which greatly increases its potential size. Lopez et al. also explore approximate querying through aggregation, which can be fairly easily
@@
 
-The introduction of an application-aware cache for query results lays the ground work for the creation of a landmark database. Such a database can store the 
+The introduction of an application-aware cache for query results lays the groundwork for the creation of a landmark database. Such a database can store the 
 locations of the highest vorticity regions in the dataset or more broadly regions of interest and their associated statistics.
@@
 
-\section{Acknowledgment}
+\section{Acknowledgments}
 The authors would like to thank the Turbulence Database Group at Johns Hopkins University as well as three anonymous reviewers for
 their insightful comments and suggestions. This work is supported in part by the National Science Foundation
-under Grants ....
+under Grants to..
 

### 2015-01-07 paper.tex
Added figure points_7rms.

@@
 We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
-the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around $10^6$ locations, which is
-sufficient to examine a time-step in detail. In the case of the vorticity these locations contain all values above $\sim$8 times the root mean square value of
-the vorticity, which is about $25\%$ of the maximum. Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ 
-of the maximum would be sufficient. 
+the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is set conservatively to $10^6$ locations, 
+which is sufficient to examine a time-step in detail. In the case of the vorticity, the values above 8 times the root mean square value, 
+which is about $25\%$ of the maximum, are contained within $2.6\times10^5$ points in each time-step. Figure \ref{fig:points_7rms} shows all the points 
+in a single time-step with values above 7 times the root mean square value. There are $2.4\times10^5$ points in the figure.
+Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ of the maximum would be sufficient. 
 At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
@@
 of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/points_7rms.png}
+\caption{Points with values above 7 times the root mean square value of the vorticity for a single time-step.}
+\label{fig:points_7rms}
+\end{figure}
 
@@
 a small set of data and do not need to perform any additional computation.
+
+Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
+locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
+reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
+easily be extended to cache the results of other query types as well if that becomes advantageous.
+
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
+space and database overhead is $\sim$40MB. 
+Therefore for a dataset containing 
+1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
+results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+require $\sim$5TB (15TB for vector fields).
 
@@
 
-Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
-locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
-reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
-easily be extended to cache the results of other query types as well if that becomes advantageous.
-
-The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
-Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
-space and database overhead is $\sim$40MB. 
-Therefore for a dataset containing 
-1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
-require $\sim$5TB (15TB for vector fields).
-
 The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
-absence of a cache.}
-\label{fig:cache_overhead}
-\end{figure}
-
 \section{Experimental Results}
@@
 database file.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
+\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
+absence of a cache.}
+\label{fig:cache_overhead}
+\end{figure}
 
@@
 
+As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
+less than 3\% and within the margin of error.
+The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
+During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
+produce a cache miss and would have to be evaluated from the raw data. 
+The execution times were averaged over 10 runs.
+We utilized 4 processes per database node for the evaluation of each query. 
+The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
+The running time increases slightly only because of the larger result set that has to be returned to the user.
+%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
+%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
+
 \begin{table}[h]
@@
 \end{table}
-
-As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
-less than 3\% and within the margin of error.
-The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
-During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
-produce a cache miss and would have to be evaluated from the raw data. 
-The execution times were averaged over 10 runs.
-We utilized 4 processes per database node for the evaluation of each query. 
-The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
-The running time increases slightly only because of the larger result set that has to be returned to the user.
-%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
-%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.
 
@@
 
+\vspace{2mm}
 \section{Related Work}
 Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
-necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array simulation
-data in the JHTDB as a collection of binary large objects in a standard relational DBMS and perform the array manipulation tasks necessary at the application level.
+necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array
+data in the JHTDB as a collection of binary large objects in a relational DBMS and perform the array manipulation tasks necessary at the application level.
 The systems that provide
@@
 dimensions with AQL and make use of the functional operators defined in AFL, such
-as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
+as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and integration with SQL:2003 syntax and semantics. It is implemented
 within the MonetDB framework \cite{MonetDB}.

### 2015-01-11 paper.tex
Added NSF grants under acknowledgments.
Minor changes and typo fixes.

@@
 %interested in out of the box. Moreover, d
-In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
+In this paper, we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
 simulation datasets stored in a cluster of relational databases. 
-Such simulations are commonplace in the study of many different physical phenomena. 
+%Such simulations are commonplace in the study of many different physical phenomena. 
 The datasets produced by these simulations are in the TB and even PB ranges. 
 Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
-locally by the user taking days or months to iterate over the entire dataset. 
+locally by the user, taking days or months to iterate over the entire dataset. 
 The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
@@
 The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
-In the case of computational turbulence this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
+Caching the results of threshold queries preserves both the I/O and computation effort used to obtain them.
+In the case of computational turbulence, this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
 the dataset as a whole, which greatly speeds up the rate of scientific exploration and discovery. 
-Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
-queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
 \end{abstract}
@@
 %% % A category with the (minimum) three required fields
-\category{H.2.8}{Database Management}{Database Applications -- scientific, spatial databases and GIS}
-%% %A category including the fourth, optional field follows...
-%% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-%\terms{Extending Database Technology}
+\category{H.2.8}{Database Management}{Database Applications -- Scientific Databases}
+\category{H.2.4}{Database Management}{Systems -- Distributed Databases}
+\category{J.2}{Computer Applications}{Physical Sciences and Engineering -- Physics}
 
@@
 
+\vspace{10pt}
 \section{Introduction}
@@
 strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
-databases. In order to reduce their running times they have to both make efficient use of the cluster resources and incorporate 
+databases. In order to reduce their running times, they have to make efficient use of the cluster resources and incorporate 
 leading data management techniques.
 
-Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
-functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
+Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest
+enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
 forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
@@
 these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
-queries in general. The established data management techniques that our approach combines make it easy to understand. It can be applied to other 
+queries in general. The established data management techniques that our approach combines make the approach easy to understand. It can be applied to other 
 scientific analysis environments, which manage large datasets in a database management system. Examples include the Sloan Digital Sky Survey
@@
 Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
-sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
+sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and applica\-tion-level
 information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
@@
 evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
-large multidimensional array datasets. Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number 
-of target locations. 
+large multidimensional array datasets. 
+Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
+Each data point appears in $N$ kernels of computation, in which $N$ is the size of the kernel.
 The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
@@
 We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
-stored in the database. The derived fields, which our method computes and operates on are derived by performing stencil kernel computations over the 
-stored array data.
+stored in the database. 
 Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
@@
 \begin{itemize}
-\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. For the first time providing large
+\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. This provides large
 data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
@@
 to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
-datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
+datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage large array datasets.
 Relational database systems however often do not support
 all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
-capabilities have to be built-in through user-defined functions or stored procedures.
+capabilities have to be built into the database through user-defined functions or stored procedures.
 
@@
 database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
-curve to distribute the data across nodes and databases. Each time-step is spatially subdivided
+curve to distribute the data across nodes and databases \cite{Perlman}. Each time-step is spatially subdivided
 into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
@@
 have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
-of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
+of the space or a collection of individual target locations within a time-step.
+
+In contrast, threshold and top-$k$ queries usually have to examine the entire 
 data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
@@
 \includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
-\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
+\caption{3D (single time-step) cut through the 4D cluster containing the most intense event.}
 \label{fig:max_vort_cluster}
@@
 can lead to new insights into the development of the most intense reconnection events of magnetic field sheets in the simulated plasma. Similarly to
-the vorticity, the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, which scientists would like to perform 
-threshold queries on certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.
+the vorticity, the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, on which scientists would like to perform 
+threshold queries certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.
 
@@
 the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
-against the specified threshold and if the value is higher it is maintained along with
+against the specified threshold and if the value is higher, it is maintained along with
 the spatial coordinates of the location in a list.
@@
 
-The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
+The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available SSD disk space, not memory.
 Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
@@
 \includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
-\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
+\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
 absence of a cache.}
@@
 The evaluation of threshold queries of derived fields from the raw data is both I/O and computationally bound. These queries examine the entire data
-volume of a simulation time-step and are therefore good candidates for a data-parallel distributed evaluation. 
+volume of a simulation time-step and are, therefore, good candidates for a data-parallel distributed evaluation. 
 Our data-parallel implementation exhibits good vertical and nearly ideal horizontal scaling as shown in Figure \ref{fig:scaling}. 
-For the scale-up (Fig. \ref{fig:scaleup}) experiments we used the same
+For the scale-up experiments (Fig. \ref{fig:scaleup}), we used the same
 queries and threshold values as for the runs shown in Figure \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
@@
 
-In all of these cases the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
+In all of these cases, the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
 time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase proportionally to the number
@@
 It is interesting to note that the time taken to perform a cache lookup is relatively small even in the case of a cache hit as can be seen in Figures 
-\ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit}, \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
+\ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit} and \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
 SSDs attached to each database node (see Fig. \ref{fig:cache_diagram}) and retrieving the data is always done through a clustered index lookup.
-In the cases of a cache hit the majority of the time is spent simply transferring the results from the database nodes to the mediator and then back to the user.
+In the cases of a cache hit, the majority of the time is spent simply transferring the results from the database nodes to the mediator and then back to the user.
 These times remain more or less constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
@@
 
-\vspace{2mm}
 \section{Related Work}
@@
 Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
-In contrast our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
+In contrast, our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
 examines the vector or scalar array data at all neighboring locations, which
@@
 
-In a system called the tree cache Lopez et al. \cite{Lopez} make use of a small application-aware cache to reduce access time to large datasets stored on
+In a system called the tree cache, Lopez et al. \cite{Lopez} make use of a small application-aware cache to reduce access time to large datasets stored on
 disk. The tree cache stores individual octants of octree datasets and exploits application-specific information to determine which octants to cache and to 
@@
 their insightful comments and suggestions. This work is supported in part by the National Science Foundation
-under Grants to..
+under Grants CDI-II: CMMI 0941530 and OCI-108849 and Johns Hopkins University's Institute for Data Intensive Engineering \& Science.
 

### 2015-01-14 paper.tex
Minor fixes addressing comments from Thomas.

@@
 examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
-%The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
+%The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
 %The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
@@
 Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
-sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and applica\-tion-level
+sufficient for some applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
 information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
@@
 %Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
-There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
-evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
-large multidimensional array datasets. 
-Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
-Each data point appears in $N$ kernels of computation, in which $N$ is the size of the kernel.
+There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. 
+The field variables have to be evaluated \emph{on-demand} from the array data stored in the database cluster. 
+The evaluations are data-intensive as they perform kernel computations on extremely large multidimensional array datasets. 
+A kernel computation computes the value at a grid location using the data points at a set of neighboring locations.
+Kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
 The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
@@
 on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
-we cache the actual results of the computation in a limited version of a semantic cache. 
+we cache query results along with query metadata and subsequent queries are evaluated against the cache.
 This leads to query performance improvement of over an order of magnitude.
@@
 
-\section{JHTDB}\label{jhtdb}
+\section{Johns Hopkins Turbulence\\ Databases}\label{jhtdb}
 Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
@@
 Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
-cached in an application-aware cache, which is a limited version of a semantic cache. 
+cached in an application-aware semantic cache. 
+In addition to query results, the cache stores their semantic descriptions and query metadata and parameters used to obtain them.
 The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.
@@
 1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
-results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
+results for threshold queries of a derived field over the entire dataset. The currently available SSD disk space per node is $\sim$200GB, which will
+be sufficient to maintain the threshold results for nearly five derived fields over the entire dataset. 
+In contrast, computing and materializing a scalar derived field for the entire dataset would
 require $\sim$5TB (15TB for vector fields).
@@
 The speedup diminishes to 1.4 times when going to 4 processes and little speedup is 
-observed with 8 processes per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
-processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. Additionally, SQL
-Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
-the boundary of each region are requested by multiple processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
+observed with 8 processes per node.
+While the computation time scales with increased process count, the time to perform I/O does not as the data on each node reside in the same 
+database table and on the same set of disks. Additionally, I/O redundancy increases as the process count increases as data along the boundary 
+of each region are requested by multiple processes.
+SQL Server already utilizes parallelism to perform the I/O even when data are retrieved utilizing a single query.
+Finally, the experiments were run on the live production database nodes, which were also servicing other user queries in addition to operating system
+and other SQL Server processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
 compared to running with a single process.
@@
 Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
-In contrast, our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
+In contrast, our approach performs a kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
 examines the vector or scalar array data at all neighboring locations, which

### 2015-01-15 paper.tex
figure 3 is for NS, not MHD
also, add "conventional" in front of theory for reconnection, since
there is an alternate theory that does predict the correct time.

@@
 
-Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
+Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the forced isotropic turbulence dataset. The locations of maximum vorticity in the dataset were clustered
 in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
@@
 is available in a database cluster, which provides built-in sophisticated analysis routines revealed flux-freezing breakdown in MHD turbulence \cite{Eyink},
-showing why solar flares last minutes rather than the millions of years that theory would predict.
+showing why solar flares last minutes rather than the millions of years that conventional theory would predict.
 

### 2015-01-15 paper.tex
Minor fixes.

--- 2015-01-14 paper.tex
@@
 it is likely that all subsequent queries to this time-step will result in a cache hit as their threshold is likely to be equal or higher than the cached one.
+Currently we observe fairly high cache-hit ratios as the workload is very structured and queries tend to examine the same regions in space and time.
 
@@
 time-steps queried. The execution times depend on the complexity of the computation needed to evaluate the particular derived field requested. Figures 
-\ref{fig:vorticity_breakdown}, \ref{fig:q_breakdown}, \ref{fig:magnetic_breakdown} show a breakdown of the execution time of threshold queries of different
+\ref{fig:vorticity_breakdown}, \ref{fig:q_breakdown} and \ref{fig:magnetic_breakdown} show a breakdown of the execution time of threshold queries of different
 derived fields, which are evaluated from the raw data and a cold cache using 4 processes per node on a 4 node cluster. Almost the entire time is spent
@@
 in the case of centralized processing is given in \cite{Ilyas}. In the case of distributed processing different approaches focus on horizontally 
-\cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deals with array data stored in
+\cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deal with array data stored in
 a relational database system. Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA
@@
 on them. The goal of both techniques is to not return large data volumes, but focus on the most intense events and interesting regions in the dataset. 
-The computation of the derived fields is carried out on the nodes of the database cluster and takes a look at the dataset as a whole, while the user obtains only
+The computation of derived fields is carried out on the nodes of the database cluster and takes a look at the dataset as a whole, while the user obtains only
 a small subset of the data, where the derived field in question is above the prescribed threshold. Sampling approaches can potentially omit some locations and

### 2015-01-15 paper.tex
Minor fixes.

@@
 
+\vspace{5pt}
 {\bf Application-aware cache for query results:} A central part of the evaluation strategy for threshold queries that we have developed is the 

### 2015-01-22 paper.tex
Added copyright notice and grant acknowledgements.

@@
 \begin{document}
+%\conferenceinfo{EDBT}{'15 Brussels, Belgium}
+%\CopyrightYear{2015} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{ISBN 978-3-89318-067-7}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
 
@@
 their insightful comments and suggestions. This work is supported in part by the National Science Foundation
-under Grants CDI-II: CMMI 0941530 and OCI-108849 and Johns Hopkins University's Institute for Data Intensive Engineering \& Science.
+under Grants CMMI-0941530, ACI-1261715, OCI-1244820 and AST-0939767 and Johns Hopkins University's 
+Institute for Data Intensive Engineering \& Science.
 
